### 2020-04-17 tidymodels/learning_tidymodels.Rmd
tutorial comparativo Caret vs Tidymodels

---
title: "Learning Tidymodels"
output: html_notebook
---

Esse notebook R tem como finalidade conhecer e exercitar o conjunto de pacotes `tidymodel`, que seria uma evolução do pacote `caret` para transformação e fitting de dados.

Reproduz ou se baseia nos seguintes tutoriais da internet:

* [Caret vs. tidymodels - comparing the old and new](https://konradsemsch.netlify.app/2019/08/caret-vs-tidymodels-comparing-the-old-and-new/)
* [IMPLEMENTING THE SUPER LEARNER WITH TIDYMODELS](https://www.alexpghayes.com/blog/implementing-the-super-learner-with-tidymodels/)

# Seguindo o exemplo

## Bibliotecas

```{r loadlib, message=FALSE, warning=FALSE}
set.seed(42)
options(max.print = 150)

library(modeldata)
library(tidymodels)
library(tidyverse)
library(caret)
library(magrittr)
library(naniar)
library(furrr)
library(skimr)
library(vip)
library(workflows)
library(tune)

plan(multicore)  
```

## Data preparation

```{r}

data("credit_data")

credit_data <- credit_data %>% set_names(tolower(names(.)))

glimpse(credit_data)

```

## Checking missing data com `naniar` e avalinado

```{r}

credit_data %>% 
  miss_var_summary()

```

## avaliando os dados

```{r}

credit_data %>% skim()

```

## Class Balance

```{r}
table(credit_data$status)
round(prop.table(table(credit_data$status)),2)
```

## modeling

```{r}

split <- rsample::initial_split(credit_data, prop = .8, strata = "status")

df_train <- training(split)
df_test  <- testing(split)

train_cv <- rsample::vfold_cv(df_train, v=5, strata = "status")
train_cv_caret <- rsample2caret(train_cv)

```

In this particular example I mainly focus on imputting missing data or assigning them a new categorical level, infrequent/ unobserved values and hot-encoding them.

```{r}

my_recipe <- df_train %>% 
  recipe(status~.) %>% 
  # imputation: add "unknown" to all missing factor values
  step_unknown(all_nominal(), -status) %>% 
  # imputation: add median to all missing numeric values
  step_medianimpute(all_numeric()) %>% 
  # compining: group factors below 5% of frequency (default) in an "infrequent_combined"
  step_other(all_nominal(), -status, other = "infrequent_combined") %>% 
  # create a "level" in the factor columns for unseen factors
  step_novel(all_nominal(), -status, new_level = "unrecorded_observation") %>% 
  # OHE
  step_dummy(all_nominal(), -status, one_hot=T)

my_recipe
my_recipe_prep <- prep(my_recipe, retain=T)

my_recipe_prep
tidy(my_recipe_prep)

```

## playing with recipe

```{r}
# some data
mydata <- tibble(
  class = as.factor(c("dog","dog","dog","cat","cat","monkey")),
  size  = c(53,42,83,20,30,70),
  weight = c(15,8,12,5,6,21)
)

# create a recipe from "mydata"
rcp <- recipe(mydata) %>% 
  # one hot encoding fo class
  step_dummy(all_nominal(), one_hot = T)

# prepare the recipe
p_rcp <- prep(rcp, verbose = T)

# apply a prepared recipe to a new data
# in this case do OHE
bake(p_rcp, mydata)

```

## Caret

```{r}

control_caret <- trainControl(
  method="cv", 
  verboseIter = F,
  classProbs = T, 
  summaryFunction = twoClassSummary,
  returnResamp = "final",
  savePredictions = "final",
  index = train_cv_caret$index,
  indexOut = train_cv_caret$indexOut
)

grid_caret <- expand.grid(
  mtry = seq(1,ncol(df_train)-1,3),
  splitrule = c("extratrees","gini"),
  min.node.size=c(1,3,5)
)

model_caret <- train(
    status~.,
    data=juice(my_recipe_prep), 
    method="ranger", 
    metric="ROC",
    trControl=control_caret, 
    tuneGrid = grid_caret,
    importance="impurity",
    num.tree=500
  )

print(model_caret)
```

```{r}

varImp(model_caret, scale=T)$importance %>% 
  rownames_to_column() %>% 
  arrange(desc(Overall))

```

```{r}

df_train_pred_caret <- model_caret$pred %>% 
  group_by(rowIndex) %>% 
  summarise(bad=mean(bad)) %>% 
  transmute(estimate=bad) %>% 
  add_column(truth=df_train$status)

# cross-validated training performance
percent(roc_auc(df_train_pred_caret, truth, estimate)$.estimate)

```
```{r}
# test performance

df_test_pred_caret <- predict(
    model_caret, 
    newdata = bake(my_recipe_prep, df_test),
    type="prob"
  ) %>% 
  as_tibble() %>% 
  transmute(estimate=bad) %>% 
  add_column(truth=df_test$status)

percent(roc_auc(df_test_pred_caret, truth, estimate)$.estimate)

```

# tidymodel

```{r}
engine_tidym <- rand_forest(
    mode = "classification",
    mtry = tune(),
    trees = tune(),
    min_n = tune()
  ) %>% 
  set_engine("ranger",importance="impurity")

print(engine_tidym)
```

```{r}
gridtm <- grid_random(
  mtry() %>% range_set(c(1, 20)),
  trees() %>% range_set(c(500, 1000)), 
  min_n() %>% range_set(c(2, 10)),
  size = 30
  )
```

```{r}

wkfl_tidym <- workflow() %>% 
  add_recipe(my_recipe) %>% 
  add_model(engine_tidym)

gridy_tidym <- tune_grid(wkfl_tidym, resamples=train_cv, grid=gridtm,
                         metric=metric_set(roc_auc),
                         control=control_grid(save_pred=T))

```

```{r}
collect_metrics(gridy_tidym)
```

```{r}

grid_tidym_best <- select_best(gridy_tidym, metric="roc_auc")
(wkfl_tidym_best <- finalize_workflow(wkfl_tidym, grid_tidym_best))
```

```{r}
(wkfl_tidym_final <- last_fit(wkfl_tidym_best, split = split))
```

```{r}
# Cross-validated training performance
percent(show_best(gridy_tidym, metric="roc_auc", n = 1)$mean)
```

```{r}
# Test performance
percent(wkfl_tidym_final$.metrics[[1]]$.estimate[[2]])
```

```{r}
vip(pull_workflow_fit(wkfl_tidym_final$.workflow[[1]]))$data
```

### 2020-04-17 tidymodels/tutorial_in_tidymodels.Rmd
tutorial de Tidymodels

---
title: "Tutorial on Tidymodel"
output: md_document
---

Reproduzindo este [artigo](https://hansjoerg.me/2020/02/09/tidymodels-for-machine-learning/) sobre `tidymodels` do [Hansjörg Plieninger](https://hansjoerg.me/page/about/)

# Tutorial on tidymodels for Machine Learning

## Setup

```{r loadlibs, message=FALSE, warning=FALSE}
library(conflicted)
library(tidymodels)
library(tidyverse)
library(ggrepel)
library(corrplot)
library(skimr)

conflict_prefer("filter","dplyr")

ggplot2::theme_set(theme_light())
```

## Dataset: `Diamonds`

```{r diamond_plot}

data("diamonds")

head(diamonds,10)

diamonds %>% 
  sample_n(2000) %>% 
  mutate_if(is.factor, as.numeric) %>% 
  select(price, everything()) %>% 
  cor() %>% 
  {.[order(abs(.[,1]), decreasing = T),
     order(abs(.[,1]), decreasing = T)]} %>% 
  corrplot(method = "number", type="upper", mar=c(0,0,1.5,0), 
           title="Correlations between price and various features of diamonds")

```

```{r}

diamonds %>% 
  skim()

```

## Separating Testing and Training Data: `rsample`

```{r}
# you know: the life, the universer... 
set.seed(42)

# training/test
dia_split <- initial_split(diamonds, prop = .1, strata = price)
dia_train <- training(dia_split)
dia_test  <- testing(dia_split)

dim(dia_train)
dim(dia_test)

# cv-folds
dia_vfold <- vfold_cv(dia_train, v=3, repeats=1, strata=price)

dia_vfold %>% 
  mutate(df_ana=map(splits, analysis),
         df_ass=map(splits, assessment)) -> folds

```

## Data Pre-Processing and Feature Engineering: `recipes`

The plot below indicates that there may be a nonlinear relationship between price and carat, and I want to address that using higher-order terms.

```{r}
qplot(carat, price, data = dia_train) +
    scale_y_continuous(trans = log_trans(), labels = function(x) round(x, -2)) +
    geom_smooth(method = "lm", formula = "y ~ poly(x, 4)") +
    labs(title = "Nonlinear relationship between price and carat of diamonds",
         subtitle = "The degree of the polynomial is a potential tuning parameter")
```

```{r}
dia_rec <- recipe(price~., data=dia_train) %>% 
  # log no preço
  step_log(all_outcomes()) %>% 
  # normializa valores (que fatoriais)
  step_normalize(all_predictors(), -all_nominal()) %>% 
  # OHE dos fatoriais
  step_dummy(all_nominal()) %>% 
  # cria uma feature (?) de polinomial
  step_poly(carat, degree = 4)

prep(dia_rec)

```

```{r}

# apply the recipe in the train data
dia_prep <- prep(dia_rec)

# extract transformed training data
dia_juiced <- juice(dia_prep)

# apply the recipe in a new data
bake(dia_prep, dia_test)

```

## Defining and Fitting Models: `parsnip`

The parsnip package has wrappers around many1 popular machine learning algorithms, and you can fit them using a unified interface. This is extremely helpful, since you have to remember only one rather then dozens of interfaces.

```{r}

# regression example
lm_model <- linear_reg() %>% 
  set_mode("regression") %>% 
  set_engine("lm")

rf_model <- rand_forest(mtry=3, trees=500, min_n=5) %>% 
  set_mode("regression") %>% 
  set_engine("ranger", importance="impurity")

```

```{r}

lm_fit <- fit(lm_model, price~., dia_juiced)
lm_fit

rf_fit <- fit(rf_model, price~., dia_juiced)
rf_fit
```

## Summarizing Fitted Models: broom

```{r}

glance(lm_fit$fit)
tidy(lm_fit$fit)
rf_fit$fit

augment(lm_fit$fit) %>% 
  rowid_to_column() %>% 
  select(rowid, price, .fitted:.std.resid)

```

### 2020-04-18 covid/acc_deads_covid_19_x_h1n1_09.R
scripts relacionados a dados da COVID-19

@@ -1,0 +1,78 @@
+library(tidyverse)
+library(glue)
+library(lubridate)
+
+## Dados: morte por influenza SINAN via InfoGripe da Fiocruz: http://info.gripe.fiocruz.br/
+
+# Scrapping dos dados da Fiocruz (InfoGripe), anos de interesse
+range <- c(2009,2016,2018,2019)
+url <- "http://info.gripe.fiocruz.br/data/resumed/3/2/{year}/0/Brasil/weekly-incidence-curve"
+
+# para cada ano, faz a requisição das mortes por "Influenza" em formato CSV e importa
+influ_scrap <- range %>% 
+  map(function(.x,.url){
+    glue(.url, year=.x)
+  }, .url=url) %>% 
+  map(read_csv, na = "null")
+
+# unificando os dataframes 
+influ <- influ_scrap %>% 
+  bind_rows() %>% 
+  # cria uma coluna de ano para cada csv
+  mutate( year = rep(range, each=57),
+          type = ifelse(year==2009|year==2016,"H1N1","Influenza")) %>% 
+  select( year, everything() )
+
+## Mortes COVD-19 notificadas via MS: https://covid.saude.gov.br/
+covid_csv <- read_delim("https://mobileapps.saude.gov.br/esus-vepi/files/unAFkcaNDeXajurGB7LChj8SgQYS2ptm/e9250e5db0e3bf01c21c407cddaa9103_Download_COVID19_20200418.csv",
+                    delim = ";")
+
+# dados vem por dia e estado
+covid <- covid_csv %>% 
+  # cria uma "semana epidemiologica"
+  mutate( epiweek = week(data)) %>% 
+  group_by(epiweek) %>% 
+  # soma os novos casos semanais
+  summarise(value = sum(obitosNovos)) %>% 
+  ungroup() %>% 
+  arrange( epiweek ) %>%  
+  # cria um "acumulado" e deixa o DF com as mesmas colunas do "influ"
+  mutate( year=2020,
+          acc_value = cumsum(value),
+          type = "Sars-Cov-2" )
+
+# para comparar, precisamos colocar os registros iniciando de uma mesma
+# semana (adjweek), para isso vamos localizar nos anos, a semana em que há
+# pelo menos X mortes registradas
+influ %>% 
+  # dados de interesse
+  select(year, epiweek, value, type) %>% 
+  group_by(year) %>% 
+  # acumula por ano
[…]

### 2020-04-18 covid/acc_srag_covid_19_x_h1n1_09.R

@@ -1,0 +1,56 @@
+library(tidyverse)
+library(glue)
+
+# Scrapping dos dados da Fiocruz (InfoGripe), anos de interesse
+range <- c(2009,2016,2018,2019,2020)
+url <- "http://info.gripe.fiocruz.br/data/detailed/1/2/{year}/52/Brasil/weekly-incidence-curve"
+
+# para cada ano, faz a requisição das notificações em formato CSV e importa
+scrap <- range %>% 
+  map(function(.x,.url){
+    glue(.url, year=.x)
+  }, .url=url) %>% 
+  map(read_csv, na = "null")
+
+# unificando os dataframes 
+cases <- scrap %>% 
+  bind_rows() %>% 
+  # cria uma coluna de ano para cada csv
+  mutate( year = rep(range, each=54) ) %>% 
+  select( year, everything() )
+
+# para comparar, precisamos colocar os registros iniciando de uma mesma
+# semana (adjweek), para isso vamos localizar nos anos, a semana em que há
+# pelo menos 500 casos registrados
+cases %>% 
+  # dados de interesse
+  select(year, epiweek, value, estimated_cases) %>% 
+  # os dados de notificação são atualizados "no passado"
+  # semanas de 2020 a partir da 14a. ainda estão com dados não confiáveis
+  # mutate( value = ifelse( year==2020&epiweek>=13,estimated_cases,value ) ) %>% 
+  group_by(year) %>% 
+  # acumula
+  mutate(acc_value=cumsum(value)) %>% 
+  nest() %>% 
+  mutate(
+    # para cada conjunto de dado por ano
+    data = map(data, function(.data){
+      .data %>% 
+        # cria uma "semana ajustada" a partir do 500o caso
+        filter(acc_value>=100) %>% 
+        mutate(adjweek = 1:nrow(.))
+    })
+  ) %>% 
+  unnest(data) %>% 
+  ungroup() %>% 
+  # ajusta o ano para fator (para plotar por cor)
+  mutate( year  = factor(year) ) %>% 
+  # pegar só o começo da epidemia
+  filter( adjweek<=25 ) %>% 
+  ggplot(aes(adjweek, acc_value, group=year, color=year)) +
+  geom_line(size=1) +
+  labs(title="Notificações de síndrome respiratória aguda grave (SRAG)",
[…]

### 2020-04-18 covid/cases_by_year.R

@@ -1,0 +1,33 @@
+library(tidyverse)
+library(glue)
+
+# Scrapping dos dados da Fiocruz (InfoGripe)
+range <- 2015:2020
+url <- "http://info.gripe.fiocruz.br/data/detailed/1/2/{year}/52/Brasil/weekly-incidence-curve"
+
+# para cada ano, faz a requisição das notificações
+scrap <- range %>% 
+  map(function(.x,.url){
+    glue(.url, year=.x)
+  }, .url=url) %>% 
+  map(read_csv, na = "null")
+
+# unificando os dataframes 
+cases <- scrap %>% 
+  bind_rows() %>% 
+  mutate( year = rep(range, each=54)) %>% 
+  select(year, everything())
+
+# plot
+cases %>% 
+  select(year, epiweek, value, estimated_cases) %>%
+  filter(year>2016) %>%  #, epiweek<=15) %>% 
+  mutate(year=factor(year)) %>% 
+  ggplot(aes(epiweek, value, group=year)) +
+  geom_line(aes(color=year), size=1) +
+  labs(title="Notificações de síndrome respiratória aguda grave (SRAG)",
+       subtitle = "fonte: www.saude.gov.br/sinan via Fiocruz",
+       x="semana do ano",
+       y="notificações SRAG") +
+  theme_minimal()
+

### 2020-04-18 covid/notif_x_proj.R

@@ -1,0 +1,14 @@
+
+
+
+cases %>% 
+  filter(year==2020, !is.na(value)) %>% 
+  select(year, epiweek, value, estimated_cases) %>% 
+  ggplot(aes(epiweek, value)) +
+  geom_line(color="darkred", size=1) +
+  geom_line(aes(epiweek, estimated_cases), linetype="dotted", size=1, alpha=.8) +
+  labs(title="Notificações de SRAG",
+       subtitle="Comparação entre notificados e projetados",
+       caption = glue("fonte: www.saude.gov.br/sinan via Fiocruz em {date}", date=Sys.Date())) +
+  theme_minimal() +
+  theme( plot.caption = element_text(hjust = 0) )

### 2020-04-18 covid/srag_covid_19_x_h1n1_09.R

@@ -1,0 +1,62 @@
+library(tidyverse)
+library(glue)
+
+# Scrapping dos dados da Fiocruz (InfoGripe), anos de interesse
+range <- c(2009,2016,2019,2020)
+url <- "http://info.gripe.fiocruz.br/data/detailed/1/2/{year}/52/Brasil/weekly-incidence-curve"
+
+# para cada ano, faz a requisição das notificações em formato CSV e importa
+scrap <- range %>% 
+  map(function(.x,.url){
+    glue(.url, year=.x)
+  }, .url=url) %>% 
+  map(read_csv, na = "null")
+
+# unificando os dataframes 
+cases <- scrap %>% 
+  bind_rows() %>% 
+  # cria uma coluna de ano para cada csv
+  mutate( year = rep(range, each=54) ) %>% 
+  select( year, everything() )
+
+# para comparar, precisamos colocar os registros iniciando de uma mesma
+# semana (adjweek), para isso vamos localizar nos anos, a semana em que há
+# pelo menos 500 casos registrados
+cases %>% 
+  # dados de interesse
+  select(year, epiweek, value, estimated_cases) %>% 
+  group_by(year) %>% 
+  nest() %>% 
+  mutate(
+    # para cada conjunto de dado por ano
+    data = map(data, function(.data){
+      .data %>% 
+        # acha a primeira semana com 500 registros
+        filter(value>=500) %>% 
+        filter(epiweek >= min(epiweek)) %>% 
+        # cria uma "semana ajustada" a partir dela
+        mutate( adjweek = 1:nrow(.) )
+    })
+  ) %>% 
+  unnest(data) %>% 
+  ungroup() %>%
+  mutate(
+    # ajusta o ano para fator (para plotar por cor)
+    year  = factor(year),
+    # os dados de notificação são atualizados "no passado"
+    # semanas de 2020 a partir da 14a. ainda estão com dados não confiáveis
+    # value = ifelse(year=="2020"&epiweek>=13,NA,value)
+  ) %>% 
+  # pegar só o começo da epidemia
+  filter( adjweek<=10 ) %>% 
+  ggplot(aes(adjweek, value, group=year, color=year)) +
+  geom_line(size=1) +
+  # plota também os dados de projeção de 2020 como pontilhado
+  geom_line(aes(adjweek, estimated_cases), linetype="dotted", size=1, color="black", alpha=.5) +
[…]

### 2020-04-18 tidymodels/tutorial_in_tidymodels.Rmd
Continuando o tutorial de Tidymodels

@@
 
-rf_fit <- fit(rf_model, price~., dia_juiced)
-rf_fit
+# rf_fit <- fit(rf_model, price~., dia_juiced)
+# rf_fit
 ```
@@
 tidy(lm_fit$fit)
-rf_fit$fit
+# rf_fit$fit
 
-augment(lm_fit$fit) %>% 
+lm_train_predicted <- augment(lm_fit$fit) %>% 
   rowid_to_column() %>% 
   select(rowid, price, .fitted:.std.resid)
+```
+
+```{r}
+
+ggplot(lm_train_predicted, aes(.fitted, price)) +
+  geom_point(alpha=.2)+
+  ggrepel::geom_label_repel(aes(label=rowid), data=filter(lm_train_predicted, abs(.resid)>2)) +
+  labs(title="Actual vs. Predicted Price of Diamonds")
 
 ```
+
+## Evaluating Model Performance: `yardstick`

### 2020-04-20 tidymodels/tutorial_in_tidymodels_1.Rmd
Tutoriais sobre Tidymodels

--- 2020-04-18 tidymodels/tutorial_in_tidymodels.Rmd
@@
 ---
-title: "Tutorial on Tidymodel"
+title: "Tutorial on Tidymodel part I"
 output: md_document
@@
 ## Evaluating Model Performance: `yardstick`
+
+We already saw performance measures RMSE and R squared in the output of glance() above.
+The yardstick package is specifically designed for such measures for both numeric and categorical outcomes, and it plays well with grouped predictions (e.g., from cross-validation).
+
+Let’s use rsample, parsnip, and yardstick for cross-validation to get a more accurate estimation of RMSE.
+
+```{r}
+
+# 3 CV fold, we can extract the "train" data and the "test/validation" data of each fold
+lm_fit2 <- dia_vfold %>% 
+  mutate(df_ana = map(splits, analysis),
+         df_ass = map(splits, assessment))
+
+lm_fit2_result <- lm_fit2 %>% 
+  mutate(
+    recipe = map(df_ana, ~prep(dia_rec, training = .x)), #prepara a receita para um dos CVs
+    df_ana = map(recipe, juice), # extrai os dados transformados na receita
+    df_ass = map2(recipe, df_ass, ~bake(.x, new_data=.y)) #aplica a receita aos dados de assessment
+  ) %>% 
+  mutate(
+    # fita um lm_model ao conjunto de dados de analise transformado
+    model_fit = map(df_ana, ~fit(lm_model, price~., data=.x)) 
+  ) %>% 
+  mutate(
+    # aplica o modelo ao dado de assessment
+    model_pred = map2(model_fit, df_ass, ~predict(.x, new_data=.y))
+  )
+
+select(lm_fit2_result, id, recipe:model_pred)
+
+```
+
+Now, we can extract the actual prices price from the assessment data and compare them to the predicted prices .pred.
+Then, the yardstick package comes into play: The function metrics() calculates three different metrics for numeric outcomes.
+Furthermore, it automatically recognizes that lm_preds is grouped by folds and thus calculates the metrics for each fold.
+
+```{r}
+lm_preds <- lm_fit2_result %>% 
+  mutate( res=map2(df_ass, model_pred, ~data.frame(price=.x$price, .pred=.y$.pred))) %>% 
+  select(id, res) %>% 
+  unnest(res) %>% 
+  group_by(id)
+
+yardstick::metrics(lm_preds, truth=price, estimate=.pred)
+```

### 2020-04-20 tidymodels/tutorial_in_tidymodels_2.Rmd
Tutoriais sobre Tidymodels

---
title: "Tutorial on Tidymodel part II"
output: md_document
---

Reproduzindo este [artigo](https://hansjoerg.me/2020/02/09/tidymodels-for-machine-learning/) sobre `tidymodels` do [Hansjörg Plieninger](https://hansjoerg.me/page/about/)

# Tutorial on tidymodels for Machine Learning

## Setup

```{r loadlibs, message=FALSE, warning=FALSE}
library(conflicted)
library(tidymodels)
library(tidyverse)
library(ggrepel)
library(corrplot)
library(skimr)

conflict_prefer("filter","dplyr")

ggplot2::theme_set(theme_light())
```

## Dataset: `Diamonds`

```{r diamond_plot}

data("diamonds")

head(diamonds,10)

# you know: the life, the universer... 
set.seed(42)

# training/test
dia_split <- initial_split(diamonds, prop = .1, strata = price)
dia_train <- training(dia_split)
dia_test  <- testing(dia_split)

dim(dia_train)
dim(dia_test)

# cv-folds
dia_vfold <- vfold_cv(dia_train, v=3, repeats=1, strata=price)

dia_vfold %>% 
  mutate(df_ana=map(splits, analysis),
         df_ass=map(splits, assessment)) -> folds

```

## Fitando uma RF

### Modelo: `parsnip`

we use `tune()` as a placeholder and let cross-validation decide on the best value for `mtry` later on.

```{r rfmodel}

# define the model
rf_model <- rand_forest(mtry = tune()) %>% #mtry will be tuned
  set_mode("regression") %>% # we'll fit for a value
  set_engine("ranger") # the RF algo

# check parameters
parameters(rf_model)

dials::mtry()

```

You can either specify the maximum for `mtry` yourself using `update()`, or you can use `finalize()` to let the data decide on the maximum.

```{r}

# setting the mtry range by hand
rf_model %>% 
  parameters() %>% 
  update(mtry=mtry(c(1L, 5L)))

# rf_model %>% 
#     parameters() %>% 
#     # Here, the maximum of mtry equals the number of predictors, i.e., 24.
#     finalize(x = select(juice(prep(dia_rec)), -price)) %>% str()
#     pull("object")

```
## Preparing data for tunning: `recipes`

As we create a feature using `step_poly()` we need to decide what are the best degree, let's `tune()` this too.

```{r}

dia_rec2 <- recipe(price~., data = dia_train) %>% 
  step_log(all_outcomes()) %>% 
  step_normalize(all_predictors(), -all_nominal()) %>% 
  step_dummy(all_nominal()) %>% 
  step_poly(carat, degree = tune())

dia_rec2 %>% 
  parameters() %>% 
  pull("object")

```

## Combine everything: `workflows`

The workflows package is designed to bundle together different parts of a machine learning pipeline like a recipe or a model.

```{r}

rf_wflow <-
    workflow() %>%
    add_model(rf_model) %>%
    add_recipe(dia_rec2)

rf_wflow

```

Second, we need to update the parameters in `rf_wflow`, because the maximum of `mtry` is not yet known and the maximum of `degree` should be four (while three is the default).

```{r}

rf_param <-
    rf_wflow %>%
    parameters() %>%
    update(mtry = mtry(range = c(3L, 5L)),
           degree = degree_int(range = c(2L, 4L)))

```

Third, we want to use cross-validation for tuning, that is, to select the best combination of the hyperparameters. Grid search will suffice. To this end, let’s create a grid of all necessary parameter combinations.

```{r}

rf_grid <- grid_regular(rf_param)
rf_grid

```

Make sure parallel processing and do a grid search

```{r}

library(future)
plan(multicore)

rf_search <- tune_grid(rf_wflow, grid = rf_grid, resamples = dia_vfold,
                       param_info = rf_param)

rf_search

```

The results can be examined using `autoplot()` and `show_best()`.

```{r}

autoplot(rf_search, metric="rmse")
show_best(rf_search, metric = "rmse",n=9) #top nine
select_best(rf_search, metric = "rmse")
select_by_one_std_err(rf_search, mtry, degree, metric="rmse")

```

## Selecting the Best Model to Make the Final Predictions

```{r}
# uses the "one-standard error rule" (Breiman _el at, 1984) 
# that selects the most simple model that is within one standard error
# of the numerically optimal results.
rf_param_final <- select_by_one_std_err(rf_search, mtry, degree, metric = "rmse")

# set 
rf_wflow_final <- finalize_workflow(rf_wflow, rf_param_final)

rf_wflow_final_fit <- fit(rf_wflow_final, data=dia_train)
rf_wflow_final_fit
```

## Predict

Now, we want to use this to `predict()` on data never seen before, namely, `dia_test`. Unfortunately, `predict(rf_wflow_final_fit, new_data = dia_test)` does not work in the present case, because the outcome is modified in the recipe via `step_log()`.

Thus, we need a little workaround: The prepped recipe is extracted from the workflow, and this can then be used to `bake()` the testing data. This baked data set together with the extracted model can then be used for the final predictions.

```{r}

# recipe from final fit
dia_rec3 <- pull_workflow_prepped_recipe(rf_wflow_final_fit)

# engine from final fit
rf_final_fit <- pull_workflow_fit(rf_wflow_final_fit)

# predicts using final fit engine with dia_test baked with final recipe
# add the y_hat to the test dataset
dia_test$.pred <- predict(rf_final_fit, new_data = bake(dia_rec3, dia_test))$.pred

# transform the Y into a log 
# apply metrics
dia_test %>%
  select(price, .pred) %>% 
  mutate(price.log = log(price)) %>% 
  metrics(truth=price.log, estimate=.pred)
```

### 2020-12-06 AdvendOfCode2020/draft_post.Rmd
Versão inicial com dataframes

---
title: "Advent Of Code 2020"
output: html_notebook
---

## Day 1: Report Repair

### Part One

After saving Christmas five years in a row, you've decided to take a vacation at a nice resort on a tropical island. Surely, Christmas will go on without you.

The tropical island has its own currency and is entirely cash-only. The gold coins used there have a little picture of a starfish; the locals just call them stars. None of the currency exchanges seem to have heard of them, but somehow, you'll need to find fifty of these coins by the time you arrive so you can pay the deposit on your room.

To save your vacation, you need to get all fifty stars by December 25th.

Collect stars by solving puzzles. Two puzzles will be made available on each day in the Advent calendar; the second puzzle is unlocked when you complete the first. Each puzzle grants one star. Good luck!

Before you leave, the Elves in accounting just need you to fix your expense report (your puzzle input); apparently, something isn't quite adding up.

Specifically, they need you to find the two entries that sum to 2020 and then multiply those two numbers together.

For example, suppose your expense report contained the following:

1721
979
366
299
675
1456

In this list, the two entries that sum to 2020 are 1721 and 299. Multiplying them together produces 1721 * 299 = 514579, so the correct answer is 514579.

Of course, your expense report is much larger. Find the two entries that sum to 2020; what do you get if you multiply them together?

```{r}
# solution using base r

# read the input as vector
input <- read.csv("day01_input.txt", header = F)[,1]

# generate all combination of 2 and generate a dataframe
comb <- as.data.frame(t(combn(input, 2)))

# sum each combinations
rSums <- rowSums(comb)

# find which one has the sum equals 2020
vals <- comb[rSums==2020,]

# multiply them
resp <- prod(unlist(vals))

resp

```

### Part two

The Elves in accounting are thankful for your help; one of them even offers you a starfish coin they had left over from a past vacation. They offer you a second one if you can find three numbers in your expense report that meet the same criteria.

Using the above example again, the three entries that sum to 2020 are 979, 366, and 675. Multiplying them together produces the answer, 241861950.

In your expense report, what is the product of the three entries that sum to 2020?

```{r}
# keeping within base r

# generic solution for any N combinations
findCombSum <- function(in.data, n.comb, match.value){

  # generate all combination of ''n.comb' and generate a dataframe
  comb <- as.data.frame(t(combn(in.data, n.comb)))

  # sum each combinations
  rSums <- rowSums(comb)

  # find which one has the sum equals 'match.value'
  vals <- comb[rSums==match.value,]

  # multiply them
  resp <- prod(unlist(vals))

  return(resp)

}

# read the input as vector
input <- read.csv("day01_input.txt", header = F)[,1]

# part 1
findCombSum(input, 2, 2020)

# part 2
findCombSum(input, 3, 2020)

```

### 2020-12-06 AdvendOfCode2020/draft_post.Rmd
First sollution puzzle 2

@@
 
-# generate all combination of 2 and generate a dataframe
-comb <- as.data.frame(t(combn(input, 2)))
+# generate all combination of 2 and generate a matrix
+comb <- combn(input, 2)
 
 # sum each combinations
-rSums <- rowSums(comb)
+sums <- colSums(comb)
 
 # find which one has the sum equals 2020
-vals <- comb[rSums==2020,]
+vals <- comb[,sums==2020]
 
 # multiply them
-resp <- prod(unlist(vals))
+resp <- prod(vals)
 
@@
 
-  # generate all combination of ''n.comb' and generate a dataframe
-  comb <- as.data.frame(t(combn(in.data, n.comb)))
+  # generate all combination of 'n.comb' and generate a matrix
+  comb <- combn(in.data, n.comb)
 
   # sum each combinations
-  rSums <- rowSums(comb)
+  sums <- colSums(comb)
 
   # find which one has the sum equals 'match.value'
-  vals <- comb[rSums==match.value,]
+  vals <- comb[,sums==match.value]
 
   # multiply them
-  resp <- prod(unlist(vals))
+  resp <- prod(vals)
 
@@
 ```
+
+## Day 2: Password Philosophy
+
+### Parte One
+
+Your flight departs in a few days from the coastal airport; the easiest way down to the coast from here is via toboggan.
+
+The shopkeeper at the North Pole Toboggan Rental Shop is having a bad day. "Something's wrong with our computers; we can't log in!" You ask if you can take a look.
+
+Their password database seems to be a little corrupted: some of the passwords wouldn't have been allowed by the Official Toboggan Corporate Policy that was in effect when they were chosen.
+
+To try to debug the problem, they have created a list (your puzzle input) of passwords (according to the corrupted database) and the corporate policy when that password was set.
+
+For example, suppose you have the following list:
+
+1-3 a: abcde
+1-3 b: cdefg
+2-9 c: ccccccccc
+
+Each line gives the password policy and then the password.
+The password policy indicates the lowest and highest number of times a given letter must appear for the password to be valid.
+For example, 1-3 a means that the password must contain a at least 1 time and at most 3 times.
+
+In the above example, 2 passwords are valid.
+The middle password, cdefg, is not; it contains no instances of b, but needs at least 1.
+The first and third passwords are valid: they contain one a or nine c, both within the limits of their respective policies.
+
+How many passwords are valid according to their policies?
+
+```{r}
+# separete and str_count come in handy for this
+library(tidyverse)
+
+dplyr:
+
+# read the data as data.
+password.check <- read.csv("./day02_input.txt", header = F) %>% 
+  set_names(c("input")) %>% 
+  separate(input, c("policy","password"), sep=": ", remove=F) %>% 
+  separate(policy, c("policy.range","policy.letter"), sep=" ") %>%
+  separate(policy.range, c("min","max"), sep = "-", convert = T) %>% 
+  mutate( count.letter = str_count(password, policy.letter)) %>% 
+  mutate( valid.password = count.letter >= min & count.letter <= max )
+
+# how many password are valid?
+sum(password.check$valid.password)
+
+```

### 2020-12-06 AdvendOfCode2020/draft_post.Rmd
day 3 part 1

@@
 ```{r}
-# separete and str_count come in handy for this
+# tidyr::separate() and stringr::str_count() come in handy for this
 library(tidyverse)
 
-dplyr:
-
-# read the data as data.
+# read the input data as data frame with a column named 'input'
 password.check <- read.csv("./day02_input.txt", header = F) %>% 
   set_names(c("input")) %>% 
+  # split the line into the password and the policy columns
   separate(input, c("policy","password"), sep=": ", remove=F) %>% 
-  separate(policy, c("policy.range","policy.letter"), sep=" ") %>%
-  separate(policy.range, c("min","max"), sep = "-", convert = T) %>% 
-  mutate( count.letter = str_count(password, policy.letter)) %>% 
-  mutate( valid.password = count.letter >= min & count.letter <= max )
+  # split the policy into range values and the letter to check
+  separate(policy, c("pol.range.min", "pol.range.max","pol.letter"),
+           sep="-| ", convert=T) %>% 
+  # count the letter in the password and check the policy range
+  mutate( count.letter = str_count(password, pol.letter),  
+          is.valid = count.letter >= pol.range.min &
+                     count.letter <= pol.range.max )
+
+# let's see what we got
+head(password.check)
 
 # how many password are valid?
-sum(password.check$valid.password)
+sum(password.check$is.valid)
 
 ```
+### Part Two
+
+While it appears you validated the passwords correctly, they don't seem to be what the Official Toboggan Corporate Authentication System is expecting.
+
+The shopkeeper suddenly realizes that he just accidentally explained the password policy rules from his old job at the sled rental place down the street!
+The Official Toboggan Corporate Policy actually works a little differently.
+
+Each policy actually describes two positions in the password, where 1 means the first character, 2 means the second character, and so on.
+(Be careful; Toboggan Corporate Policies have no concept of "index zero"!) Exactly one of these positions must contain the given letter.
+Other occurrences of the letter are irrelevant for the purposes of policy enforcement.
+
+Given the same example list from above:
+
+1-3 a: abcde is valid: position 1 contains a and position 3 does not.
+1-3 b: cdefg is invalid: neither position 1 nor position 3 contains b.
+2-9 c: ccccccccc is invalid: both position 2 and position 9 contain c.
+
+How many passwords are valid according to the new interpretation of the policies?
+
+```{r}
+# tidyr::separate() and stringr::str_count() come in handy for this
+library(tidyverse)
+
+# read the input data as data frame with a column named 'input'
+password.check <- read.csv("./day02_input.txt", header = F) %>% 
+  set_names(c("input")) %>% 
+  # split the line into the password and the policy columns
+  separate(input, c("policy","password"), sep=": ", remove=F) %>% 
+  # split the policy into range values and the letter to check
+  separate(policy, c("pol.range.min", "pol.range.max","pol.letter"),
+           sep="-| ", convert=T) %>% 
+  # part 1 
+  # count the letter in the password and check the policy range
+  mutate( count.letter = str_count(password, pol.letter),  
+          is.valid.part1 = count.letter >= pol.range.min &
+                           count.letter <= pol.range.max ) %>% 
+  # part 2
+  # get the letters in positions 'min' and 'max"' and "XOR" check
+  mutate(
+    letter.at.pos1 = str_sub(password, pol.range.min, pol.range.min),
+    letter.at.pos2 = str_sub(password, pol.range.max, pol.range.max),
+    is.valid.part2 = xor(letter.at.pos1==pol.letter, letter.at.pos2==pol.letter)
+  )
+
+# part one: how many password are valid (repeat letter policy)?
+sum(password.check$is.valid.part1)
+
+# part tow: how many password are valid (letter at position xor policy?
+sum(password.check$is.valid.part2)
+
+```
+## Day 3: Toboggan Trajectory
+
+### Part 1
+
+With the toboggan login problems resolved, you set off toward the airport.
+While travel by toboggan might be easy, it's certainly not safe: there's very minimal steering and the area is covered in trees.
+You'll need to see which angles will take you near the fewest trees.
+
+Due to the local geology, trees in this area only grow on exact integer coordinates in a grid. You make a map (your puzzle input) of the open squares (.) and trees (#) you can see. For example:
+
+```
+..##.......
+#...#...#..
+.#....#..#.
+..#.#...#.#
+.#...##..#.
+..#.##.....
+.#.#.#....#
+.#........#
+#.##...#...
+#...##....#
+.#..#...#.#
+```
+
+These aren't the only trees, though; due to something you read about once involving arboreal genetics and biome stability, the same pattern repeats to the right many times:
+
+```
+..##.........##.........##.........##.........##.........##.......  --->
+#...#...#..#...#...#..#...#...#..#...#...#..#...#...#..#...#...#..
+.#....#..#..#....#..#..#....#..#..#....#..#..#....#..#..#....#..#.
+..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#
+.#...##..#..#...##..#..#...##..#..#...##..#..#...##..#..#...##..#.
+..#.##.......#.##.......#.##.......#.##.......#.##.......#.##.....  --->
+.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#
+.#........#.#........#.#........#.#........#.#........#.#........#
+#.##...#...#.##...#...#.##...#...#.##...#...#.##...#...#.##...#...
+#...##....##...##....##...##....##...##....##...##....##...##....#
+.#..#...#.#.#..#...#.#.#..#...#.#.#..#...#.#.#..#...#.#.#..#...#.#  --->
+```
+
+You start on the open square (.) in the top-left corner and need to reach the bottom (below the bottom-most row on your map).
+
+The toboggan can only follow a few specific slopes (you opted for a cheaper model that prefers rational numbers); start by counting all the trees you would encounter for the slope right 3, down 1:
+
+From your starting position at the top-left, check the position that is right 3 and down 1.
+Then, check the position that is right 3 and down 1 from there, and so on until you go past the bottom of the map.
+
+The locations you'd check in the above example are marked here with O where there was an open square and X where there was a tree:
+
+```
+..##.........##.........##.........##.........##.........##.......  --->
+#..O#...#..#...#...#..#...#...#..#...#...#..#...#...#..#...#...#..
+.#....X..#..#....#..#..#....#..#..#....#..#..#....#..#..#....#..#.
+..#.#...#O#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#
+.#...##..#..X...##..#..#...##..#..#...##..#..#...##..#..#...##..#.
+..#.##.......#.X#.......#.##.......#.##.......#.##.......#.##.....  --->
+.#.#.#....#.#.#.#.O..#.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#
+.#........#.#........X.#........#.#........#.#........#.#........#
+#.##...#...#.##...#...#.X#...#...#.##...#...#.##...#...#.##...#...
+#...##....##...##....##...#X....##...##....##...##....##...##....#
+.#..#...#.#.#..#...#.#.#..#...X.#.#..#...#.#.#..#...#.#.#..#...#.#  --->
+```
+
+In this example, traversing the map using this slope would cause you to encounter 7 trees.
+
+Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?
+
+```{r}
+
+# read the files as vector of string
+input <- read.csv("./day03_input.txt", header = F)[,1]
+
+# create a "char" matrix of "." and "#"
+char_map <- strsplit(input, "") %>% 
+  unlist() %>% 
+  matrix(nrow = length(input), byrow = T)
+
+# convert in to 0 and 1 (1 for the trees) 
+int_map <- 1*(char_map == "#")
+
+# lets subset the matrix with the trajectory
+
+# the hill trajectory: 3 to the left and 1 to the bottom
+shifts <- c(
+  0,1,
+  0,2,
+  0,3,
+  1,3
+) %>% matrix(nrow=4, byrow=T)
+
+# starting from top left
+base_pos <- c(1,1)
+trajectory <- base_pos
+
+# until the last row
+for(i in 1:(nrow(int_map)-1)){
+  steps <- rep(base_pos,4) %>% 
+    matrix(nrow = 4, byrow = T) +
+    shifts
+
+  base_pos <- steps[4,]
+
+  trajectory <- rbind(trajectory, base_pos)
+}
+
+# the columns repeat adjusting
+y_coords <- trajectory[,2] %% ncol(int_map)
+y_coords[y_coords==0] <- ncol(int_map)
+
+trajectory[,2] <- y_coords
+
+sum(int_map[trajectory])
+
+```
+
+### Part 2
+
+Time to check the rest of the slopes - you need to minimize the probability of a sudden arboreal stop, after all.
+
+Determine the number of trees you would encounter if, for each of the following slopes, you start at the top-left corner and traverse the map all the way to the bottom:
+
+Right 1, down 1.
+Right 3, down 1. (This is the slope you already checked.)
+Right 5, down 1.
+Right 7, down 1.
+Right 1, down 2.
+In the above example, these slopes would find 2, 7, 3, 4, and 2 tree(s) respectively; multiplied together, these produce the answer 336.
+
+What do you get if you multiply together the number of trees encountered on each of the listed slopes?
+
+```{r}
+# read the files as vector of string
+input <- read.csv("./day03_input.txt", header = F)[,1]
+
+# create a "char" matrix of "." and "#"
+char_map <- strsplit(input, "") %>% 
+  unlist() %>% 
+  matrix(nrow = length(input), byrow = T)
+
+# convert in to 0 and 1 (1 for the trees) 
+int_map <- 1*(char_map == "#")
+
+# lets subset the matrix with the trajectory
+
+# the hill trajectory: 3 to the left and 1 to the bottom
+shifts <- c(
+  0,1,
+  0,2,
+  0,3,
+  1,3
+) %>% matrix(nrow=4, byrow=T)
+
+# starting from top left
+base_pos <- c(1,1)
+trajectory <- base_pos
+
+# until the last row
+for(i in 1:(nrow(int_map)-1)){
+  steps <- rep(base_pos,4) %>% 
+    matrix(nrow = 4, byrow = T) +
+    shifts
+
+  base_pos <- steps[4,]
+
+  trajectory <- rbind(trajectory, base_pos)
+}
+
+# the columns repeat adjusting
+y_coords <- trajectory[,2] %% ncol(int_map)
+y_coords[y_coords==0] <- ncol(int_map)
+
+trajectory[,2] <- y_coords
+
+sum(int_map[trajectory])
+
+```

### 2020-12-06 AdvendOfCode2020/draft_post.Rmd
part 2 of day 3 final

@@
 ```{r}
+# count trees in a trajectory
+# receives a int matrix where "1" is a tree and "0" not
+# receives a "shift pattern" for each step down
+countTreesByShift <- function(shift.pattern, area.map){
+
+  # starting from top left
+  base_pos <- c(1,1)
+  trajectory <- base_pos
+
+  # apply the pattern until the bottom
+  for(i in 1:(nrow(area.map)-1)){
+    step <- base_pos + shift.pattern
+    base_pos <- step
+    trajectory <- rbind(trajectory, step)
+  }
+
+  # the columns repeat adjusting the bound
+  y_coords <- trajectory[,2] %% ncol(area.map)
+  y_coords[y_coords==0] <- ncol(area.map)
+  trajectory[,2] <- y_coords
+  # avoid out of index in the number of rows
+  trajectory <- trajectory[trajectory[,1]<=nrow(area.map),]
+
+  # subset the map with trajectory coordinates and sum
+  return(sum(int_map[trajectory]))
+}
+
 # read the files as vector of string
@@
 
-# lets subset the matrix with the trajectory
+# test the patterns
 
-# the hill trajectory: 3 to the left and 1 to the bottom
-shifts <- c(
-  0,1,
-  0,2,
-  0,3,
-  1,3
-) %>% matrix(nrow=4, byrow=T)
+# part 1
+countTreesByShift(c(1,3),int_map)
 
-# starting from top left
-base_pos <- c(1,1)
-trajectory <- base_pos
-
-# until the last row
-for(i in 1:(nrow(int_map)-1)){
-  steps <- rep(base_pos,4) %>% 
-    matrix(nrow = 4, byrow = T) +
-    shifts
-
-  base_pos <- steps[4,]
-
-  trajectory <- rbind(trajectory, base_pos)
-}
-
-# the columns repeat adjusting
-y_coords <- trajectory[,2] %% ncol(int_map)
-y_coords[y_coords==0] <- ncol(int_map)
-
-trajectory[,2] <- y_coords
-
-sum(int_map[trajectory])
+# part 2
+c(1,1,1,3,1,5,1,7,2,1) %>% 
+  matrix(ncol=2, byrow = T) %>%
+  split(1:nrow(.)) %>% 
+  purrr::map_dbl(countTreesByShift, area.map=int_map) %>% 
+  prod()
 

### 2020-12-06 AdvendOfCode2020/draft_post.Rmd
day 04 part 1

@@
 ```
+## Day 4: Passport Processing
+
+### Part One
+
+You arrive at the airport only to realize that you grabbed your North Pole Credentials instead of your passport.
+While these documents are extremely similar, North Pole Credentials aren't issued by a country and therefore aren't actually valid documentation for travel in most of the world.
+
+It seems like you're not the only one having problems, though; a very long line has formed for the automatic passport scanners, and the delay could upset your travel itinerary.
+
+Due to some questionable network security, you realize you might be able to solve both of these problems at the same time.
+
+The automatic passport scanners are slow because they're having trouble detecting which passports have all required fields. The expected fields are as follows:
+
+byr (Birth Year)
+iyr (Issue Year)
+eyr (Expiration Year)
+hgt (Height)
+hcl (Hair Color)
+ecl (Eye Color)
+pid (Passport ID)
+cid (Country ID)
+
+Passport data is validated in batch files (your puzzle input). Each passport is represented as a sequence of key:value pairs separated by spaces or newlines. Passports are separated by blank lines.
+
+Here is an example batch file containing four passports:
+
+ecl:gry pid:860033327 eyr:2020 hcl:#fffffd
+byr:1937 iyr:2017 cid:147 hgt:183cm
+
+iyr:2013 ecl:amb cid:350 eyr:2023 pid:028048884
+hcl:#cfa07d byr:1929
+
+hcl:#ae17e1 iyr:2013
+eyr:2024
+ecl:brn pid:760753108 byr:1931
+hgt:179cm
+
+hcl:#cfa07d eyr:2025 pid:166559648
+iyr:2011 ecl:brn hgt:59in
+
+The first passport is valid - all eight fields are present. The second passport is invalid - it is missing hgt (the Height field).
+
+The third passport is interesting; the only missing field is cid, so it looks like data from North Pole Credentials, not a passport at all!
+Surely, nobody would mind if you made the system temporarily ignore missing cid fields.
+Treat this "passport" as valid.
+
+The fourth passport is missing two fields, cid and byr. Missing cid is fine, but missing any other field is not, so this passport is invalid.
+
+According to the above rules, your improved system would report 2 valid passports.
+
+Count the number of valid passports - those that have all required fields. Treat cid as optional. In your batch file, how many passports are valid?
+
+```{r}
+library(tidyverse)
+
+# read data as text (vector of string by lines)
+input <- read_lines("./day04_input.txt")
+
+# change a empty line to "|" (trick to split later)
+input[input==""] <- "|"
+
+# colapse the data and split the data 
+# into a list of fields x value
+passports <- input %>% 
+  # colapse into one string
+  str_c(collapse = " ") %>% 
+  # split each passport data
+  str_split("\\|") %>% 
+  unlist() %>% 
+  # for each passport split the fields
+  str_split(" ") %>% 
+  # pre-process the fields (spliting )
+  map(function(line){
+    line[line!=""] %>% 
+      str_split(":", simplify = T) %>% 
+      return()
+  })
+
+is.valid.passdata <- function(pass.data){
+  fields <- pass.data[,1]
+  # has all fields
+  if (length(fields) == 8) return(T)
+  # has all fields but cid
+  if (length(fields)==7 & length(fields[fields=="cid"])==0 ) return(T)
+  return(F)
+}
+
+passports %>% 
+  map_lgl(is.valid.passdata) %>% 
+  sum()
+
+```

### 2020-12-08 AdvendOfCode2020/draft_post.Rmd
Advent of Code 2020 Day 04 Part 2

@@
 ```
+
+### Part two
+
+The line is moving more quickly now, but you overhear airport security talking about how passports with invalid data are getting through. Better add some data validation, quick!
+
+You can continue to ignore the cid field, but each other field has strict rules about what values are valid for automatic validation:
+
+```
+byr (Birth Year) - four digits; at least 1920 and at most 2002.
+iyr (Issue Year) - four digits; at least 2010 and at most 2020.
+eyr (Expiration Year) - four digits; at least 2020 and at most 2030.
+hgt (Height) - a number followed by either cm or in:
+ - If cm, the number must be at least 150 and at most 193.
+ - If in, the number must be at least 59 and at most 76.
+hcl (Hair Color) - a # followed by exactly six characters 0-9 or a-f.
+ecl (Eye Color) - exactly one of: amb blu brn gry grn hzl oth.
+pid (Passport ID) - a nine-digit number, including leading zeroes.
+cid (Country ID) - ignored, missing or not.
+```
+
+Your job is to count the passports where all required fields are both present and valid according to the above rules. Here are some example values:
+
+```
+byr valid:   2002
+byr invalid: 2003
+
+hgt valid:   60in
+hgt valid:   190cm
+hgt invalid: 190in
+hgt invalid: 190
+
+hcl valid:   #123abc
+hcl invalid: #123abz
+hcl invalid: 123abc
+
+ecl valid:   brn
+ecl invalid: wat
+
+pid valid:   000000001
+pid invalid: 0123456789
+```
+
+Here are some invalid passports:
+
+```
+eyr:1972 cid:100
+hcl:#18171d ecl:amb hgt:170 pid:186cm iyr:2018 byr:1926
+
+iyr:2019
+hcl:#602927 eyr:1967 hgt:170cm
+ecl:grn pid:012533040 byr:1946
+
+hcl:dab227 iyr:2012
+ecl:brn hgt:182cm pid:021572410 eyr:2020 byr:1992 cid:277
+
+hgt:59cm ecl:zzz
+eyr:2038 hcl:74454a iyr:2023
+pid:3556412378 byr:2007
+```
+
+Here are some valid passports:
+
+pid:087499704 hgt:74in ecl:grn iyr:2012 eyr:2030 byr:1980
+hcl:#623a2f
+
+eyr:2029 ecl:blu cid:129 byr:1989
+iyr:2014 pid:896056539 hcl:#a97842 hgt:165cm
+
+hcl:#888785
+hgt:164cm byr:2001 iyr:2015 cid:88
+pid:545766238 ecl:hzl
+eyr:2022
+
+iyr:2010 hgt:158cm hcl:#b6652a ecl:blu byr:1944 eyr:2021 pid:093154719
+
+Count the number of valid passports - those that have all required fields and valid values. Continue to treat cid as optional. In your batch file, how many passports are valid?
+
+```{r}
+library(tidyverse)
+
+# read data as text (vector of string by lines)
+input <- read_lines("day04_input.txt")
+
+# change a empty line to "|" (trick to split later)
+input[input==""] <- "|"
+
+# colapse the data and split the data 
+# into a list of fields x value
+passports <- input %>% 
+  # colapse into one string
+  str_c(collapse = " ") %>% 
+  # split each passport data
+  str_split("\\|") %>% 
+  unlist() %>% 
+  # for each passport split the fields
+  str_split(" ") %>% 
+  # pre-process the fields
+  map(function(line){
+    # remove empty strings and split into "name":"value
+    pd <- str_split(line[line!=""], ":", simplify = T)
+    # return a list of fields
+    pd[,2] %>% 
+      split(1:length(.)) %>% 
+      set_names(pd[,1]) %>% 
+      as_tibble() %>% 
+      return()
+  })
+
+# puts in a tibble
+pass.check <- tibble(
+    pdata = passports
+  ) %>% 
+  # unnest
+  unnest(pdata) %>% 
+  # validates each field
+  # byr (Birth Year) - four digits; at least 1920 and at most 2002.
+  mutate( byr = as.integer(byr) ) %>% 
+  mutate( chk.byr = (!is.na(byr) & byr>=1920 & byr<=2002) ) %>% 
+  # iyr (Issue Year) - four digits; at least 2010 and at most 2020.
+  mutate( iyr = as.integer(iyr) ) %>% 
+  mutate( chk.iyr = (!is.na(iyr) & iyr>=2010 & iyr<=2020) ) %>% 
+  # eyr (Expiration Year) - four digits; at least 2020 and at most 2030.
+  mutate( eyr = as.integer(eyr) ) %>% 
+  mutate( chk.eyr = (!is.na(eyr) & eyr>=2020 & eyr<=2030)) %>% 
+  # hgt (Height) - a number followed by either cm or in:
+  mutate( hgt.fmt.chk = (!is.na(hgt) & str_detect(hgt,"[0-9]+(in|cm)")) ) %>% 
+  # separates value of unit to check the ranges
+  mutate( hgt.val = as.integer(str_extract(hgt,"[0-9]+")),
+          hgt.unit = str_extract(hgt,"(in|cm)")) %>% 
+  mutate( chk.hgt = case_when(
+    # invalid format or empty field
+    !hgt.fmt.chk ~ F,
+    # - If cm, the number must be at least 150 and at most 193.
+    hgt.unit == "cm" ~ (hgt.val>=150 & hgt.val<=193),
+    # - If in, the number must be at least 59 and at most 76.
+    hgt.unit == "in" ~ (hgt.val>=59 & hgt.val<=76),
+    T ~ F # otherwise?
+  )) %>% 
+  # hcl (Hair Color) - a # followed by exactly six characters 0-9 or a-f.
+  mutate( chk.hcl = (!is.na(hcl) & str_detect(hcl, "#[0-9a-f]{6}") & nchar(hcl)==7) ) %>% 
+  # ecl (Eye Color) - exactly one of: amb blu brn gry grn hzl oth.
+  mutate( chk.ecl = (!is.na(ecl) & ecl %in% c("amb","blu","brn","gry","grn","hzl","oth"))) %>% 
+  # pid (Passport ID) - a nine-digit number, including leading zeroes.
+  mutate( chk.pid = (!is.na(pid) & str_detect(pid,"[0-9]{9}") * nchar(pid)==9) ) %>%
+  # cid (Country ID) - ignored, missing or not. 
+  # check all policies
+  mutate( is.valid = (chk.byr & chk.iyr & chk.eyr & chk.hgt & chk.hcl & chk.ecl & chk.pid))
+
+# how many is valid?
+pass.check %>% 
+  count(is.valid)
+
+```

### 2020-12-08 AdvendOfCode2020/passport_check.R

@@ -1,0 +1,156 @@
+library(tidyverse)
+
+# read data as text (vector of string by lines)
+input <- read_lines("AdvendOfCode2020/day04_input.txt")
+
+# change a empty line to "|" (trick to split later)
+input[input==""] <- "|"
+
+# colapse the data and split the data 
+# into a list of fields x value
+passports <- input %>% 
+  # colapse into one string
+  str_c(collapse = " ") %>% 
+  # split each passport data
+  str_split("\\|") %>% 
+  unlist() %>% 
+  # for each passport split the fields
+  str_split(" ") %>% 
+  # pre-process the fields
+  map(function(line){
+    # remove empty strings and split into "name":"value
+    pd <- str_split(line[line!=""], ":", simplify = T)
+    # return a list of fields
+    pd[,2] %>% 
+      split(1:length(.)) %>% 
+      set_names(pd[,1]) %>% 
+      return()
+  })
+
+
+is.valid.passdata <- function(pdata){
+  
+  # # it hasn't enough fields
+  # if (length(pdata)<7) return(F)
+  # 
+  # # it hasn't enough field
+  # if ( length(pdata)==7 & is.null(pdata[["cid"]]) ) return(F)
+  
+  # byr (Birth Year) - four digits; at least 1920 and at most 2002.
+  byr <- (
+    !is.null(pdata[["byr"]]) &&
+      nchar(pdata[["byr"]])==4 &&
+      as.integer(pdata[["byr"]]) >= 1920 &&
+      as.integer(pdata[["byr"]]) <= 2002 
+  )
+  
+  # iyr (Issue Year) - four digits; at least 2010 and at most 2020.
+  iyr <- (
+    !is.null(pdata[["iyr"]]) &&
+      nchar(pdata[["iyr"]])==4 &&
+      as.integer(pdata[["iyr"]]) >= 2010 &&
+      as.integer(pdata[["iyr"]]) <= 2020 
+  )
+  
+  # eyr (Expiration Year) - four digits; at least 2020 and at most 2030.
+  eyr <- (
+    !is.null(pdata[["eyr"]]) &&
+      nchar(pdata[["eyr"]])==4 &&
+      as.integer(pdata[["eyr"]]) >= 2020 &&
[…]

### 2020-12-08 AdvendOfCode2020/draft_post_2.Rmd
Day 6 parte 1 e 2

---
title: "Advent of Code | Day 5-8"
output: html_notebook
---

## Day 5: Binary Boarding

### Part 1

You board your plane only to discover a new problem: you dropped your boarding pass! You aren't sure which seat is yours, and all of the flight attendants are busy with the flood of people that suddenly made it through passport control.

You write a quick program to use your phone's camera to scan all of the nearby boarding passes (your puzzle input); perhaps you can find your seat through process of elimination.

Instead of zones or groups, this airline uses binary space partitioning to seat people. A seat might be specified like `FBFBBFFRLR`, where `F` means "`front`", `B` means "`back`", `L` means "`left`", and `R` means "`right`".

The `first 7` characters will either be `F` or `B`; these specify `exactly one of the 128 rows on the plane` (numbered 0 through 127). `Each letter tells you which half of a region the given seat is in`. Start with the whole list of rows; the first letter indicates whether the seat is in the front (0 through 63) or the back (64 through 127). The next letter indicates which half of that region the seat is in, and so on until you're left with exactly one row.

For example, consider just the first seven characters of `FBFBBFFRLR`:

```
Start by considering the whole range, rows 0 through 127.
F means to take the lower half, keeping rows 0 through 63.
B means to take the upper half, keeping rows 32 through 63.
F means to take the lower half, keeping rows 32 through 47.
B means to take the upper half, keeping rows 40 through 47.
B keeps rows 44 through 47.
F keeps rows 44 through 45.
The final F keeps the lower of the two, row 44.
```

The `last three` characters will be either `L` or `R`; these specify exactly `one of the 8 columns of seats` on the plane (numbered 0 through 7). The same process as above proceeds again, this time with only three steps. `L` means to keep `the lower half`, while `R` means to keep `the upper half`.

For example, consider just the last 3 characters of `FBFBBFFRLR`:

```
Start by considering the whole range, columns 0 through 7.
R means to take the upper half, keeping columns 4 through 7.
L means to take the lower half, keeping columns 4 through 5.
The final R keeps the upper of the two, column 5.
So, decoding FBFBBFFRLR reveals that it is the seat at row 44, column 5.
```

Every seat also has a `unique seat ID`: multiply the row by 8, then add the column. In this example, the seat has `ID 44 * 8 + 5 = 357`.

Here are some other boarding passes:

```
BFFFBBFRRR: row 70, column 7, seat ID 567.
FFFBBBFRRR: row 14, column 7, seat ID 119.
BBFFBBFRLL: row 102, column 4, seat ID 820.
```

As a sanity check, look through your list of boarding passes. What is the highest seat ID on a boarding pass?

```{r}
code <- "BFFFBBFRRR"
seat.code <- unlist(strsplit(code,""))
seat.row.code <- seat.code[1:7]
seat.col.code <- seat.code[8:10]

row.map <- 0:127
for(step in seat.row.code){
  print(step)
  lower.map <- row.map[1:(length(row.map)/2)]
  high.map  <- row.map[((length(row.map)/2)+1):length(row.map)]
  if(step=="F") {
    row.map <- lower.map 
  } else {
    row.map <- high.map 
  }
  print(row.map)
}

row.number <- row.map

col.map <- 0:7
for(step in seat.col.code){

  lower.map <- col.map[1:(length(col.map)/2)]
  high.map  <- col.map[((length(col.map)/2)+1):length(col.map)]
  if(step=="L") {
    col.map <- lower.map 
  } else {
    col.map <- high.map 
  }
  print(col.map)
}

col.number <- col.map

# ID: multiply the row by 8, then add the column
row.number*8+col.number
```
```{r}

findInMap <- function(steps, pos.map, lower.code){

  for (step in steps){

    if(step==lower.code){
      index <- 1:(length(pos.map)/2)
    } else {
      index <- ((length(pos.map)/2)+1):length(pos.map)
    }

    pos.map <- pos.map[index]

  }

  return(pos.map)
}

calcSeatId <- function(code){
  seat.code <- unlist(strsplit(code,""))
  seat.row.code <- seat.code[1:7]
  seat.col.code <- seat.code[8:10]

  rnum <- findInMap(seat.row.code, 0:127, "F")
  cnum <- findInMap(seat.col.code, 0:7, "L")

  rnum*8+cnum
}

c("FBFBBFFRLR", "BFFFBBFRRR", "FFFBBBFRRR", "BBFFBBFRLL")

sapply(c("FBFBBFFRLR", "BFFFBBFRRR", "FFFBBBFRRR", "BBFFBBFRLL"),calcSeatId)

input <- readr::read_lines("day05_input.txt")
ids <- sapply(input, calcSeatId)
max(ids)

```

### Part Two

Ding! The "fasten seat belt" signs have turned on. Time to find your seat.

It's a completely full flight, so your seat should be the only missing boarding pass in your list. However, there's a catch: some of the seats at the very front and back of the plane don't exist on this aircraft, so they'll be missing from your list as well.

Your seat wasn't at the very front or back, though; the seats with IDs +1 and -1 from yours will be in your list.

What is the ID of your seat?

```{r}

min.seat <- min(ids)
max.seat <- max(ids)
full.seats <- min.seat:max.seat

setdiff(full.seats, ids)

```
## Day 6: Custom Customs

### Part One

As your flight approaches the regional airport where you'll switch to a much larger plane, customs declaration forms are distributed to the passengers.

The form asks a series of 26 yes-or-no questions marked a through z. All you need to do is identify the questions for which anyone in your group answers "yes". Since your group is just you, this doesn't take very long.

However, the person sitting next to you seems to be experiencing a language barrier and asks if you can help. For each of the people in their group, you write down the questions for which they answer "yes", one per line. For example:

```
abcx
abcy
abcz
```

In this group, there are 6 questions to which anyone answered "yes": a, b, c, x, y, and z. (Duplicate answers to the same question don't count extra; each question counts at most once.)

Another group asks for your help, then another, and eventually you've collected answers from every group on the plane (your puzzle input). Each group's answers are separated by a blank line, and within each group, each person's answers are on a single line. For example:

```
abc

a
b
c

ab
ac

a
a
a
a

b
```

This list represents answers from five groups:

```
The first group contains one person who answered "yes" to 3 questions: a, b, and c.
The second group contains three people; combined, they answered "yes" to 3 questions: a, b, and c.
The third group contains two people; combined, they answered "yes" to 3 questions: a, b, and c.
The fourth group contains four people; combined, they answered "yes" to only 1 question, a.
The last group contains one person who answered "yes" to only 1 question, b.
In this example, the sum of these counts is 3 + 3 + 3 + 1 + 1 = 11.
```

For each group, count the number of questions to which anyone answered "yes". What is the sum of those counts?

```{r}
library(stringr)
library(magrittr)
library(purrr)

input <- readr::read_lines("day06_input.txt")

input[input==""] <- "|"

input %>% 
  str_c(collapse = "") %>% 
  str_split("\\|") %>% 
  unlist() %>% 
  map_int(function(.x){
    .x %>% 
      strsplit("") %>% 
      unlist() %>% 
      unique() %>% 
      length()
  }) %>% 
  sum()

```

### Part Two

As you finish the last group's customs declaration, you notice that you misread one word in the instructions:

You don't need to identify the questions to which anyone answered "yes"; you need to identify the questions to which everyone answered "yes"!

Using the same example as above:

```
abc

a
b
c

ab
ac

a
a
a
a

b
```

This list represents answers from five groups:

```
In the first group, everyone (all 1 person) answered "yes" to 3 questions: a, b, and c.
In the second group, there is no question to which everyone answered "yes".
In the third group, everyone answered yes to only 1 question, a. Since some people did not answer "yes" to b or c, they don't count.
In the fourth group, everyone answered yes to only 1 question, a.
In the fifth group, everyone (all 1 person) answered "yes" to 1 question, b.
In this example, the sum of these counts is 3 + 0 + 1 + 1 + 1 = 6.
```

For each group, count the number of questions to which everyone answered "yes". What is the sum of those counts?

```{r}
library(stringr)
library(magrittr)
library(purrr)

input <- readr::read_lines("day06_input.txt")

input[input==""] = "|"

in.data <- input %>% 
  str_c(collapse = ">") %>% 
  str_split("\\|") %>% 
  unlist() %>% 
  map(function(.x){
    resp <- unlist(str_split(.x, ">"))
    return(resp[resp!=""])
  })

tibble(input=in.data) %>% 
  mutate( gcount = map(input, groupCount)) %>% 
  mutate( g.id = row_number() ) %>% 
  unnest( gcount ) %>% 
  count( g.id, reply ) %>% 
  summarise( resp = sum(n) )

```

### 2020-12-08 AdvendOfCode2020/group_counting.R

@@ -1,0 +1,28 @@
+library(tidyverse)
+
+gresps
+
+groupCount <- function(g.resps){
+  tb <- tibble(
+    group_replies = list(g.resps)
+  ) %>% 
+    mutate( n.resps = map_int(group_replies, length)) %>% 
+    mutate( prep.resp = map(group_replies, function(iresp){
+      iresp %>% 
+        strsplit("") %>% 
+        unlist()
+    })) %>% 
+    mutate(
+      count.resp = list(map_df(prep.resp,function(resps){
+        tibble(reply=resps)
+      })) )
+  
+  tb %>% 
+    mutate(  count.resp = map(count.resp, ~count(.x, reply)) ) %>% 
+    unnest(count.resp) %>% 
+    filter(n==n.resps) %>% 
+    select(reply, n, n.resps) %>% 
+    return()
+}
+
+groupCount(c("abc"))  

### 2020-12-11 AdvendOfCode2020/draft_post_2.Rmd
Day 07 part 1

@@
 
+groupCount <- function(g.resps){
+  tb <- tibble(
+    group_replies = list(g.resps)
+  ) %>% 
+    mutate( n.resps = map_int(group_replies, length)) %>% 
+    mutate( prep.resp = map(group_replies, function(iresp){
+      iresp %>% 
+        strsplit("") %>% 
+        unlist()
+    })) %>% 
+    mutate(
+      count.resp = list(map_df(prep.resp,function(resps){
+        tibble(reply=resps)
+      })) )
+
+  tb %>% 
+    mutate(  count.resp = map(count.resp, ~count(.x, reply)) ) %>% 
+    unnest(count.resp) %>% 
+    filter(n==n.resps) %>% 
+    select(reply, n, n.resps) %>% 
+    return()
+}
+
 input <- readr::read_lines("day06_input.txt")

### 2020-12-11 AdvendOfCode2020/draft_post_3.Rmd
Day 07 part 1

---
title: "R Notebook"
output: html_notebook
---

## Day 7: Handy Haversacks

You land at the regional airport in time for your next flight. In fact, it looks like you'll even have time to grab some food: all flights are currently delayed due to issues in luggage processing.

Due to recent aviation regulations, many rules (your puzzle input) are being enforced about bags and their contents; bags must be color-coded and must contain specific quantities of other color-coded bags. Apparently, nobody responsible for these regulations considered how long they would take to enforce!

For example, consider the following rules:

```
light red bags contain 1 bright white bag, 2 muted yellow bags.
dark orange bags contain 3 bright white bags, 4 muted yellow bags.
bright white bags contain 1 shiny gold bag.
muted yellow bags contain 2 shiny gold bags, 9 faded blue bags.
shiny gold bags contain 1 dark olive bag, 2 vibrant plum bags.
dark olive bags contain 3 faded blue bags, 4 dotted black bags.
vibrant plum bags contain 5 faded blue bags, 6 dotted black bags.
faded blue bags contain no other bags.
dotted black bags contain no other bags.
```

These rules specify the required contents for 9 bag types. In this example, every faded blue bag is empty, every vibrant plum bag contains 11 bags (5 faded blue and 6 dotted black), and so on.

You have a shiny gold bag. If you wanted to carry it in at least one other bag, how many different bag colors would be valid for the outermost bag? (In other words: how many colors can, eventually, contain at least one shiny gold bag?)

In the above rules, the following options would be available to you:

A bright white bag, which can hold your shiny gold bag directly.
A muted yellow bag, which can hold your shiny gold bag directly, plus some other bags.
A dark orange bag, which can hold bright white and muted yellow bags, either of which could then hold your shiny gold bag.
A light red bag, which can hold bright white and muted yellow bags, either of which could then hold your shiny gold bag.

So, in this example, the number of bag colors that can eventually contain at least one shiny gold bag is 4.

How many bag colors can eventually contain at least one shiny gold bag? (The list of rules is quite long; make sure you get all of it.)

To begin, get your puzzle input

```{r}
library(tidyverse)

# reads the input as a vector of strings
input <- readLines("./day07_input.txt")

# transforms the strings in a data frame with the rules
bag.rules <- tibble(input) %>% 
  # separates the bag from the content rule
  separate(input, into=c("bag","contains"), sep=" contain ") %>% 
  # for each back split the rule in different types of bag
  # we build a 'tidy' dataset, each line maps from a bag to a content
  group_by(bag) %>% 
  nest() %>% 
  ungroup() %>% 
  mutate( contains =  map(data,~unlist(str_split(.x,", "))) ) %>% 
  unnest(contains) %>% 
  select(-data) %>% 
  # now we splt the bag type and color 
  separate(bag, into = c("bag.type","bag.color"), sep = " ", extra = "drop") %>%
  # also for the content rule
  separate(contains, into = c("ctn.bag.qtd","ctn.bag.type","ctn.bag.color"), sep=" ", extra = "drop") %>% 
  # some adjust to bags with no content
  mutate( ctn.bag.qtd = if_else(ctn.bag.qtd=="no","0",ctn.bag.qtd) ) %>% 
  mutate( ctn.bag.qtd = as.integer(ctn.bag.qtd), 
          ctn.bag.type = if_else( ctn.bag.qtd==0, as.character(NA), ctn.bag.type),
          ctn.bag.color = if_else( ctn.bag.qtd==0, as.character(NA), ctn.bag.color) )

# let's see what we have
bag.rules

library(tidygraph)

# we build a graph, first create a "from->to" edge list
bag.graph <- bag.rules %>% 
  transmute( from = paste0(ctn.bag.type,"_",ctn.bag.color),
             to = paste0(bag.type,"_",bag.color),
             weight = ctn.bag.qtd ) %>% 
  # remove bags with no content
  filter(weight!=0) %>% 
  # keep the capacity (as weight) and directed
  as_tbl_graph(bag.graph, directed = T)

# what we have?
bag.graph

# we query the graph asking for all paths from the "shiny_old" node
bag.graph %>% 
  all_simple_paths(from="shiny_gold") %>% 
  # get only the "end" node of each path
  map_chr(~names(tail(.x,1))) %>% 
  unique() %>% 
  # counts it
  length()

```
--- Part Two ---

It's getting pretty expensive to fly these days - not because of ticket prices, but because of the ridiculous number of bags you need to buy!

Consider again your shiny gold bag and the rules from the above example:

```
faded blue bags contain 0 other bags.
dotted black bags contain 0 other bags.
vibrant plum bags contain 11 other bags: 5 faded blue bags and 6 dotted black bags.
dark olive bags contain 7 other bags: 3 faded blue bags and 4 dotted black bags.
```

So, a single shiny gold bag must contain 1 dark olive bag (and the 7 bags within it) plus 2 vibrant plum bags (and the 11 bags within each of those): 1 + 1*7 + 2 + 2*11 = 32 bags!

Of course, the actual rules have a small chance of going several levels deeper than this example; be sure to count all of the bags, even if the nesting becomes topologically impractical!

Here's another example:

```
shiny gold bags contain 2 dark red bags.
dark red bags contain 2 dark orange bags.
dark orange bags contain 2 dark yellow bags.
dark yellow bags contain 2 dark green bags.
dark green bags contain 2 dark blue bags.
dark blue bags contain 2 dark violet bags.
dark violet bags contain no other bags.
```

In this example, a single shiny gold bag must contain 126 other bags.

How many individual bags are required inside your single shiny gold bag?

### 2020-12-11 AdvendOfCode2020/draft_post_3.Rmd
Day 07 part 2

@@
 How many individual bags are required inside your single shiny gold bag?
+
+```{r}
+
+# reads the input as a vector of strings
+input <- readLines("./day07_input.txt")
+
+# transforms the strings in a data frame with the rules
+bag.rules <- tibble(input) %>% 
+  # separates the bag from the content rule
+  separate(input, into=c("bag","contains"), sep=" contain ") %>% 
+  # for each back split the rule in different types of bag
+  # we build a 'tidy' dataset, each line maps from a bag to a content
+  group_by(bag) %>% 
+  nest() %>% 
+  ungroup() %>% 
+  mutate( contains =  map(data,~unlist(str_split(.x,", "))) ) %>% 
+  unnest(contains) %>% 
+  select(-data) %>% 
+  # now we splt the bag type and color 
+  separate(bag, into = c("bag.type","bag.color"), sep = " ", extra = "drop") %>%
+  # also for the content rule
+  separate(contains, into = c("ctn.bag.qtd","ctn.bag.type","ctn.bag.color"), sep=" ", extra = "drop") %>% 
+  # some adjust to bags with no content
+  mutate( ctn.bag.qtd = if_else(ctn.bag.qtd=="no","0",ctn.bag.qtd) ) %>% 
+  mutate( ctn.bag.qtd = as.integer(ctn.bag.qtd), 
+          ctn.bag.type = if_else( ctn.bag.qtd==0, as.character(NA), ctn.bag.type),
+          ctn.bag.color = if_else( ctn.bag.qtd==0, as.character(NA), ctn.bag.color) )
+
+# transforms the bag rules into a graph from bag to contains
+bag.graph <- bag.rules %>% 
+  transmute( from = paste0(bag.type,"_",bag.color),
+             to = paste0(ctn.bag.type,"_",ctn.bag.color),
+             weight = ctn.bag.qtd ) %>% 
+  filter(weight!=0) %>% 
+  # keep the capacity (as weight) and directed
+  as_tbl_graph(bag.graph, directed = T)
+
+# finds all paths starting 
+paths <- bag.graph %>% 
+  all_simple_paths(from = "shiny_gold")
+
+# for each path gets the weight between the nodes
+paths %>% 
+  map_dbl(function(.p, .bg){
+    # run through the path
+    weights <- vector(mode="double", (length(.p)-1))
+    for (i in 1:(length(.p)-1)){
+      # gets the weights along the pagh
+      weights[i] <- .bg %>% 
+        filter( from == .p[i], 
+                to   == .p[i+1]) %>% 
+        pull(weight)
+    }
+    # return the product of the weights
+    return(prod(weights))
+  }, .bg=activate(bag.graph, edges)) %>% 
+  # sum them up
+  sum()
+
+```

### 2020-12-13 AdvendOfCode2020/day_08.Rmd
Resolução dos puzzle do dia 8 datasets

---
title: "R Notebook"
output: html_notebook
---

## Day 8: Handheld Halting

### Part One

Your flight to the major airline hub reaches cruising altitude without incident. While you consider checking the in-flight menu for one of those drinks that come with a little umbrella, you are interrupted by the kid sitting next to you.

Their handheld game console won't turn on! They ask if you can take a look.

You narrow the problem down to a strange infinite loop in the boot code (your puzzle input) of the device. You should be able to fix it, but first you need to be able to run the code in isolation.

The boot code is represented as a text file with one instruction per line of text. Each instruction consists of an operation (`acc`, `jmp`, or `nop`) and an argument (a signed number like `+4` or `-20`).

`acc` increases or decreases a single global value called the accumulator by the value given in the argument. For example, `acc +7` would increase the accumulator by `7`. *The accumulator starts at 0*. After an `acc` instruction, the instruction immediately below it is executed next.
`jmp` jumps to a new instruction relative to itself. The next instruction to execute is found using the argument as an offset from the `jmp` instruction; for example, `jmp +2` would skip the next instruction, `jmp +1` would continue to the instruction immediately below it, and `jmp -20` would cause the instruction 20 lines above to be executed next.
`nop` stands for No OPeration - it does nothing. The instruction immediately below it is executed next.
For example, consider the following program:

```
nop +0
acc +1
jmp +4
acc +3
jmp -3
acc -99
acc +1
jmp -4
acc +6
```

These instructions are visited in this order:

```
nop +0  | 1
acc +1  | 2, 8(!)
jmp +4  | 3
acc +3  | 6
jmp -3  | 7
acc -99 |
acc +1  | 4
jmp -4  | 5
acc +6  |
```

First, the `nop +0` does nothing. Then, the accumulator is increased from 0 to 1 (`acc +1`) and `jmp +4` sets the next instruction to the other `acc +1` near the bottom. After it increases the accumulator from 1 to 2, `jmp -4` executes, setting the next instruction to the only `acc +3`. It sets the accumulator to 5, and `jmp -3` causes the program to continue back at the first `acc +1`.

This is an infinite loop: with this sequence of jumps, the program will run forever. The moment the program tries to run any instruction a second time, you know it will never terminate.

Immediately before the program would run an instruction a second time, the value in the accumulator is 5.

Run your copy of the boot code. Immediately before any instruction is executed a second time, what value is in the accumulator?

```{r warning=FALSE, message=FALSE}
library(tidyverse)

input <- readLines("./day08_test_01.txt")

decodeProgram <- function(.input) {
  tibble(input=.input) %>% 
    separate(input, into = c("cmd","arg"), sep=" ", convert = T) %>% 
    mutate( exec = 0 ) %>% 
    return()
}

program <- decodeProgram(input)

# start conditions
acc <- 0
ptr <- 1

# run the program until before a command be executed a second time
while( i<= nrow(program) && program[ptr, ]$exec!=1 ){

    # get the command and arg
  cmd <- program[ptr,]$cmd
  arg <- program[ptr,]$arg

  # update acc
  acc <- case_when(
    cmd=="acc" ~ acc + arg,
    T ~ acc
  )

  # mark command execution
  program[ptr,]$exec <- 1

  # update the pointer
  ptr <- case_when(
    cmd=="jmp" ~ ptr+arg,
    T ~ ptr + 1
  )

}

# what are the accumulator?
acc
```

### Part Two

After some careful analysis, you believe that exactly one instruction is corrupted.

Somewhere in the program, either a `jmp` is supposed to be a `nop`, or a `nop` is supposed to be a `jmp.` (No `acc` instructions were harmed in the corruption of this boot code.)

*The program is supposed to terminate by attempting to execute an instruction immediately after the last instruction in the file*. By changing exactly one `jmp` or `nop`, you can repair the boot code and make it terminate correctly.

For example, consider the same program from above:

```
nop +0
acc +1
jmp +4
acc +3
jmp -3
acc -99
acc +1
jmp -4
acc +6
```

If you change the first instruction from `nop +0` to `jmp +0`, it would create a single-instruction infinite loop, never leaving that instruction. If you change almost any of the `jmp` instructions, the program will still eventually find another `jmp` instruction and loop forever.

However, if you change the second-to-last instruction (from `jmp -4` to `nop -4`), the program terminates! The instructions are visited in this order:

```
nop +0  | 1
acc +1  | 2
jmp +4  | 3
acc +3  |
jmp -3  |
acc -99 |
acc +1  | 4
nop -4  | 5
acc +6  | 6
```

After the last instruction (`acc +6`), the program terminates by attempting to run the instruction below the last instruction in the file. With this change, after the program terminates, the accumulator contains the value 8 (`acc +`1, `acc +1`, `acc +6`).

Fix the program so that it terminates normally by changing exactly one `jmp` (to `nop`) or `nop` (to `jmp`). What is the value of the accumulator after the program terminates?

```{r}

executeProgram <- function(.program, acc=0, ptr=1, canChange=F){

  # run the program until before a command be executed a second time
  while( ptr > 0 &&
         ptr <= nrow(.program) &&
         .program[ptr, ]$exec!=1 ) {

      # get the command and arg
    cmd <- .program[ptr,]$cmd
    arg <- .program[ptr,]$arg

    # update acc
    acc <- case_when(
      cmd=="acc" ~ acc + arg,
      T ~ acc
    )

    # mark command execution
    .program[ptr,]$exec <- 1

    # update the pointer
    ptr <- case_when(
      cmd=="jmp" ~ ptr+arg,
      T ~ ptr + 1
    )
  }

  # verify the exit state
  result <- case_when(
    ptr < 1                      ~ "error",
    ptr <= nrow(.program)       ~ "loop",
    ptr == (nrow(.program) + 1) ~ "end",
    T ~"error"
  )

  # return
  return(list(exit=result, acc=acc))

}

input <- readLines("./day08_input.txt")
program <- decodeProgram(input)
result <- executeProgram(program)
changeIndex <- which(program$cmd %in% c("jmp","nop"))
i <- 1

while( result$exit!="end" && i <= length(changeIndex) ){

  step <- program[changeIndex[i],]
  chg.program <- program

  chg.program[changeIndex[i],1] <-  case_when(
    step$cmd=="jmp" ~ "nop",
    step$cmd=="nop" && step$arg!=0 ~ "jmp",
    T ~ chg.program[changeIndex[i],]$cmd
  )

  result <- executeProgram(chg.program)
  i <- i +1
}

result

```

### 2020-12-13 AdvendOfCode2020/draft_post_3.Rmd
Resolução dos puzzle do dia 8 datasets

@@
 ---
+
+## Graph Generic Test
+
+```{r}
+library(tidygraph)
+library(tidyverse)
+library(ggraph)
+
+g <- tibble(
+  from = LETTERS[sample(1:8,12, replace = T)],
+  to   = LETTERS[sample(1:8,12, replace = T)],
+  dist = sample(1:8,12, replace = T)
+) %>% 
+  as_tbl_graph()
+
+plotGraph <- function(g){
+  g %>% 
+    ggraph(layout = "kk") +
+      geom_edge_fan(alpha=0.3) +
+      geom_node_point(color="blue",alpha=0.8, size=8) +
+      geom_node_text(aes(label=name), color="white") +
+      theme_graph()
+}
+
+plotGraph(g)
+
+g %N>%
+  filter(name %in% c("G","H","C")) %>% 
+  plotGraph()
+
+g %>% 
+  to_subgraph(name %in% c("G","H","C")) %>%
+  .[[1]] %>% 
+  class()
+
+```
 
@@
 
+```
 A bright white bag, which can hold your shiny gold bag directly.
@@
 A light red bag, which can hold bright white and muted yellow bags, either of which could then hold your shiny gold bag.
+```
 
@@
 # reads the input as a vector of strings
-input <- readLines("./day07_input.txt")
+input <- readLines("./day07_test_01.txt")
 
-# transforms the strings in a data frame with the rules
-bag.rules <- tibble(input) %>% 
+# function to decode the string bag rules in a tibble
+decodeBagRules <- function(.input){
+  tibble(input=.input) %>% 
   # separates the bag from the content rule
@@
           ctn.bag.color = if_else( ctn.bag.qtd==0, as.character(NA), ctn.bag.color) )
+}
+
+# transforms the strings in a data frame with the rules
+bag.rules <- decodeBagRules(input)
 
@@
 
+plotGraph(bag.graph)
+
 # we query the graph asking for all paths from the "shiny_old" node
@@
   length()
+
+bag.subgraphs <- bag.graph %>% 
+  all_simple_paths(from="shiny_gold") %>% 
+  map(function(.p,.g){
+    to_subgraph(.g, name %in% names(.p)) %>% 
+      .[[1]] %E>% 
+      as_tibble() %>% 
+      pull(weight) %>% 
+      sum()
+  },.g=bag.graph)
+
+bag.subgraphs %>% 
+  map(function(.sg){
+    class(.sg)
+  })
 

### 2020-12-13 AdvendOfCode2020/day09.Rmd
Puzzle 01 Day 09 Test Data

---
title: "R Notebook"
output: html_notebook
---

## Day 9: Encoding Error

### Parte One

With your neighbor happily enjoying their video game, you turn your attention to an open data port on the little screen in the seat in front of you.

Though the port is non-standard, you manage to connect it to your computer through the clever use of several paperclips. Upon connection, the port outputs a series of numbers (your puzzle input).

The data appears to be encrypted with the eXchange-Masking Addition System (XMAS) which, conveniently for you, is an old cypher with an important weakness.

XMAS starts by transmitting a preamble of 25 numbers. After that, each number you receive should be the sum of any two of the 25 immediately previous numbers. The two numbers will have different values, and there might be more than one such pair.

For example, suppose your preamble consists of the numbers 1 through 25 in a random order. To be valid, the next number must be the sum of two of those numbers:

26 would be a valid next number, as it could be 1 plus 25 (or many other pairs, like 2 and 24).
49 would be a valid next number, as it is the sum of 24 and 25.
100 would not be valid; no two of the previous 25 numbers sum to 100.
50 would also not be valid; although 25 appears in the previous 25 numbers, the two numbers in the pair must be different.

Suppose the 26th number is 45, and the first number (no longer an option, as it is more than 25 numbers ago) was 20. Now, for the next number to be valid, there needs to be some pair of numbers among 1-19, 21-25, or 45 that add up to it:

26 would still be a valid next number, as 1 and 25 are still within the previous 25 numbers.
65 would not be valid, as no two of the available numbers sum to it.
64 and 66 would both be valid, as they are the result of 19+45 and 21+45 respectively.

Here is a larger example which only considers the previous 5 numbers (and has a preamble of length 5):

35
20
15
25
47
40
62
55
65
95
102
117
150
182
127
219
299
277
309
576

In this example, after the 5-number preamble, almost every number is the sum of two of the previous 5 numbers; the only number that does not follow this rule is 127.

The first step of attacking the weakness in the XMAS data is to find the first number in the list (after the preamble) which is not the sum of two of the 25 numbers before it. What is the first number that does not have this property?

```{r}

library(purrr)
library(magrittr)

input <- as.double(readLines("./day09_test_01.txt"))

checkData <- function(.preamble, .number) {
  return(.number %in% colSums(combn(.preamble,2)))
}

pre.size <- 5

numChecked <- (pre.size+1):length(input) %>% 
  map_lgl(function(.i, .d, .ps){
    checkData(.d[(.i-.ps):(.i-1)], .d[.i]) %>% 
      return()
  }, .d=input, .ps=pre.size)

input[min(pre.size + which(numChecked==FALSE))]
```

### 2020-12-13 AdvendOfCode2020/day09.Rmd
Puzzle 2 Day 09

@@
 
-XMAS starts by transmitting a preamble of 25 numbers.
-After that, each number you receive should be the sum of any two of the 25 immediately previous numbers.
+XMAS starts by transmitting a *preamble of 25 numbers*.
+After that, each number you receive *should be the sum of any two of the 25 immediately previous numbers*.
 The two numbers will have different values, and there might be more than one such pair.
@@
 
+```
 26 would be a valid next number, as it could be 1 plus 25 (or many other pairs, like 2 and 24).
@@
 50 would also not be valid; although 25 appears in the previous 25 numbers, the two numbers in the pair must be different.
+```
 
@@
 
+```
 26 would still be a valid next number, as 1 and 25 are still within the previous 25 numbers.
@@
 64 and 66 would both be valid, as they are the result of 19+45 and 21+45 respectively.
+```
 
@@
 
+```
 35
@@
 576
+```
 
-In this example, after the 5-number preamble, almost every number is the sum of two of the previous 5 numbers; the only number that does not follow this rule is 127.
+In this example, after the 5-number preamble, almost every number is the sum of two of the previous 5 numbers; the only number that does not follow this rule is *127*.
 
@@
 
-input <- as.double(readLines("./day09_test_01.txt"))
+input <- as.double(readLines("./day09_input.txt"))
 
@@
 
-pre.size <- 5
+pre.size <- 25
 
@@
 ```
+
+### Part Two
+
+The final step in breaking the XMAS encryption relies on the invalid number you just found: *you must find a contiguous set of at least two numbers in your list which sum to the invalid number from step 1*.
+
+Again consider the above example:
+
+```
+35
+20
+15
+25
+47
+40
+62
+55
+65
+95
+102
+117
+150
+182
+127
+219
+299
+277
+309
+576
+```
+
+In this list, adding up all of the numbers from 15 through 40 produces the invalid number from step 1, 127. (Of course, the contiguous set of numbers in your actual list might be much longer.)
+
+To find the encryption weakness, add together the smallest and largest number in this contiguous range; in this example, these are 15 and 47, producing 62.
+
+What is the encryption weakness in your XMAS-encrypted list of numbers?
+
+```{r}
+
+library(purrr)
+library(magrittr)
+
+input <- as.double(readLines("./day09_input.txt"))
+
+checkData <- function(.preamble, .number) {
+  return(.number %in% colSums(combn(.preamble,2)))
+}
+
+pre.size <- 25
+
+numChecked <- (pre.size+1):length(input) %>% 
+  map_lgl(function(.i, .d, .ps){
+    checkData(.d[(.i-.ps):(.i-1)], .d[.i]) %>% 
+      return()
+  }, .d=input, .ps=pre.size)
+
+step1 <- input[min(pre.size + which(numChecked==FALSE))]
+
+rangeComb <- combn(1:length(input),2)
+
+rngChecked <- rangeComb %>% 
+  apply(2,function(.range, .data, .numCheck){
+    return(sum(.data[.range[1]:.range[2]])==.numCheck)
+  }, .data=input, .numCheck=step1)
+
+answerRange <- rangeComb[,which(rngChecked==T)]
+contRange <- input[answerRange[1]:answerRange[2]]
+
+resp <- min(contRange) + max(contRange)
+
+resp
+
+```

### 2020-12-13 AdvendOfCode2020/day09.Rmd
Day 09 | commented code

@@
 ```{r}
-
 library(purrr)
@@
 
+# reads the input data as an array of numbers
 input <- as.double(readLines("./day09_input.txt"))
 
+# function that checks if a number is the sum
+# of all combinations of two from the preamble
 checkData <- function(.preamble, .number) {
@@
 
+# preamble size
 pre.size <- 25
 
-numChecked <- (pre.size+1):length(input) %>% 
+# checks, along the data it there is a sum in the preamble
+# numCheck return an array of result check for all data
+numChecked <- (pre.size+1):length(input) %>% # the range of the data (input-preamble)
   map_lgl(function(.i, .d, .ps){
+    # calc the preamble and the number to be checked
     checkData(.d[(.i-.ps):(.i-1)], .d[.i]) %>% 
@@
 
+# once checked all number, get the first index that fail and
+# returns that input value at this position
 input[min(pre.size + which(numChecked==FALSE))]
@@
 ```{r}
-
-library(purrr)
 library(magrittr)
 
+# reads the input data as an array of numbers
 input <- as.double(readLines("./day09_input.txt"))
 
+# function that checks if a number is the sum
+# of all combinations of two from the preamble
 checkData <- function(.preamble, .number) {
@@
 
+# once checked all number, get the first index that fail and
+# returns that input value at this position
+# store the answer in the 'step1' var to be used in sequence
 step1 <- input[min(pre.size + which(numChecked==FALSE))]
 
+# Now we must find a continuous range of falues which the sum is equal
+# the value found in step 1
+
+# Calc all possible "continuous" range index
 rangeComb <- combn(1:length(input),2)
 
+# for each, check if the sum of its values match the target value
+# we test all then and store the test result
 rngChecked <- rangeComb %>% 
@@
 
+# get the index range which the range matched the sum
 answerRange <- rangeComb[,which(rngChecked==T)]
+
+# get the values in this range
 contRange <- input[answerRange[1]:answerRange[2]]
 
+# sum the smallest and the largest value in this range
 resp <- min(contRange) + max(contRange)
 
+# this is our answer
 resp

### 2020-12-13 AdvendOfCode2020/day_08.Rmd
Commented code from day 08

@@
 
+# Parses the array of string into a dataframe with columns "cmd" and "arg"
 decodeProgram <- function(.input) {
@@
 
-  # run the program until before a command be executed a second time
-  while( ptr > 0 &&
-         ptr <= nrow(.program) &&
-         .program[ptr, ]$exec!=1 ) {
+  # run the program until..
+  while( ptr > 0 &&                # ptr jumps outside the bordering -> its a error
+         ptr <= nrow(.program) &&  # ptr jumps outside the bordering -> its a error or finishes
+         .program[ptr, ]$exec!=1 ) { # ptr points to a command already executed -> its a loop
 
@@
 
+# read the input as an array of string
 input <- readLines("./day08_input.txt")
+
+# parses the input data
 program <- decodeProgram(input)
+
+# execut3 the program as is
 result <- executeProgram(program)
+
+# locate the possible changes in the program
 changeIndex <- which(program$cmd %in% c("jmp","nop"))
@@
 
-while( result$exit!="end" && i <= length(changeIndex) ){
+# for each one, change the command and test
+while( result$exit!="end" &&        # until we found a change that works
+       i <= length(changeIndex) ){  # until the end of possibilities
 
+  # get an step to be changed
   step <- program[changeIndex[i],]
+
+  # modify the program
   chg.program <- program 
-
   chg.program[changeIndex[i],1] <-  case_when(
-    step$cmd=="jmp" ~ "nop",
-    step$cmd=="nop" && step$arg!=0 ~ "jmp",
-    T ~ chg.program[changeIndex[i],]$cmd
+    step$cmd=="jmp" ~ "nop",                # change 'jmp' to 'nop'
+    step$cmd=="nop" && step$arg!=0 ~ "jmp", # change 'nop' to 'jmp' (when args is not zero: loop)
+    T ~ chg.program[changeIndex[i],]$cmd    # not change anything
   )
 
+  # execute the changed program
   result <- executeProgram(chg.program)
+
+  # next change...
   i <- i +1
@@
 
+# check the result
 result

### 2020-12-22 AdvendOfCode2020/day09.Rmd
Day 11 puzzle 1

@@
 
+#### Solution
+
+The solutions is pretty straighforward, we use the preamble as a _slide window_ along the transmission data and do a combination of two from the preamble and sum it, so we check if the next number is one of the possible value from it.
+
 ```{r}
@@
 
+#### Solution
+
+We'll apply the same idea from part 1, to find which value is invalid.
+After that we create a array of continuous values from one to size of transmission, combine then, sum and find which ones are equal to the invalid value found.
+
 ```{r}
@@
 
-# Now we must find a continuous range of falues which the sum is equal
+# Now we must find a continuous range of values which the sum is equal
 # the value found in step 1
 
-# Calc all possible "continuous" range index
+# Calculates all possible "continuous" range index
 rangeComb <- combn(1:length(input),2)
@@
 ```
+
+There is it!

### 2020-12-22 AdvendOfCode2020/day_08.Rmd
Day 11 puzzle 1

@@
 Run your copy of the boot code. Immediately before any instruction is executed a second time, what value is in the accumulator?
+
+#### Solution
+
+This is fun remembers me my old assembly classes in the college.
+
+We'll create a function to decode the input text in to a data frame of commands and arguments and a execution counter, to check if that command was already executed and so detect a loop.
+And we'll _run_ the data frame processing each command (row) until the end it or to find a loop.
+We update the execution counter for each row executed.
 
@@
 
+#### Solution
+
+We'll use the same idea to part one, but here let's put the code that _runs_ a program (the code data frame) in a more sophisticated function.
+This function will returns if a program ended if with a error (jump to a invalid position), with a loop (executed the same instruction twice) or with success (run the last command).
+
+To find which command we need to change, we just find with positions of instructions `nop` or `jmp` in the original code, and for each one, change it, run the changed program and check the outcome, until we find a programa that works.
+
 ```{r}
 
+# Emulates the execution of a code returning:  
+# ERROR if the execution jumps to a invalid position
+# LOOP, if executes the same instruction twice
+# END if run the last command nicelly
 executeProgram <- function(.program, acc=0, ptr=1, canChange=F){
@@
 ```
+That is our response.

### 2020-12-22 AdvendOfCode2020/day_10.Rmd
Day 11 puzzle 1

---
title: "R Notebook"
output: html_notebook
---

## Day 10: Adapter Array

### Part One

Patched into the aircraft's data port, you discover weather forecasts of a massive tropical storm. Before you can figure out whether it will impact your vacation plans, however, your device suddenly turns off!

Its battery is dead.

You'll need to plug it in. There's only one problem: the charging outlet near your seat produces the wrong number of jolts. Always prepared, you make a list of all of the joltage adapters in your bag.

Each of your joltage adapters is rated for a specific output joltage (your puzzle input). Any given adapter can take an input 1, 2, or 3 jolts lower than its rating and still produce its rated output joltage.

In addition, your device has a built-in joltage adapter rated for 3 jolts higher than the highest-rated adapter in your bag. (If your adapter list were 3, 9, and 6, your device's built-in adapter would be rated for 12 jolts.)

Treat the charging outlet near your seat as having an effective joltage rating of 0.

Since you have some time to kill, you might as well test all of your adapters. Wouldn't want to get to your resort and realize you can't even charge your device!

If you use every adapter in your bag at once, what is the distribution of joltage differences between the charging outlet, the adapters, and your device?

For example, suppose that in your bag, you have adapters with the following joltage ratings:

```
16
10
15
5
1
11
7
19
6
12
4
```

With these adapters, your device's built-in joltage adapter would be rated for 19 + 3 = 22 jolts, 3 higher than the highest-rated adapter.

Because adapters can only connect to a source 1-3 jolts lower than its rating, in order to use every adapter, you'd need to choose them like this:

* The charging outlet has an effective rating of 0 jolts, so the only adapters that could connect to it directly would need to have a joltage rating of 1, 2, or 3 jolts. Of these, only one you have is an adapter rated 1 jolt (difference of 1).
* From your 1-jolt rated adapter, the only choice is your 4-jolt rated adapter (difference of 3).
* From the 4-jolt rated adapter, the adapters rated 5, 6, or 7 are valid choices. However, in order to not skip any adapters, you have to pick the adapter rated 5 jolts (difference of 1).
* Similarly, the next choices would need to be the adapter rated 6 and then the adapter rated 7 (with difference of 1 and 1).
* The only adapter that works with the 7-jolt rated adapter is the one rated 10 jolts (difference of 3).
* From 10, the choices are 11 or 12; choose 11 (difference of 1) and then 12 (difference of 1).
* After 12, only valid adapter has a rating of 15 (difference of 3), then 16 (difference of 1), then 19 (difference of 3).
* Finally, your device's built-in adapter is always 3 higher than the highest adapter, so its rating is 22 jolts (always a difference of 3).

In this example, when using every adapter, there are 7 differences of 1 jolt and 5 differences of 3 jolts.

Here is a larger example:

```
28
33
18
42
31
14
46
20
48
47
24
23
49
45
19
38
39
11
1
32
25
35
8
17
7
9
4
2
34
10
3
```

In this larger example, in a chain that uses all of the adapters, there are 22 differences of 1 jolt and 10 differences of 3 jolts.

Find a chain that uses all of your adapters to connect the charging outlet to your device's built-in adapter and count the joltage differences between the charging outlet, the adapters, and your device. What is the number of 1-jolt differences multiplied by the number of 3-jolt differences?

#### Solution

This is simple, we sort a sequence of adapters, calculate the difference between then and just count the differences of 1 and 3 jolts.

```{r}

# finds the # of differences of 1 and 3 jolts
countJoltageDiff <- function(.input){
  # adds the outlet and the device joltages and sort it
  joltages <- sort(c(0, input, max(.input)+3))

  # calc the differences
  jolt_diffs <- diff(joltages)

  # count joltages diff 1 and joltages diff 3
  return(list("1"=sum(jolt_diffs==1),"3"=sum(jolt_diffs==3)))
}

# reads the input as a array of integers
input <- as.integer(readLines("./day10_test_01.txt"))
countJoltageDiff(input)

input <- as.integer(readLines("./day10_test_02.txt"))
countJoltageDiff(input)

input <- input <- as.integer(readLines("./day10_input.txt"))
jdiffs <- countJoltageDiff(input)

# response (# of diffs 1 * # of diffs 3)
jdiffs$`1` * jdiffs$`3`

```
### Part Two

To completely determine whether you have enough adapters, you'll need to figure out how many different ways they can be arranged. Every arrangement needs to connect the charging outlet to your device. The previous rules about when adapters can successfully connect still apply.

The first example above (the one that starts with 16, 10, 15) supports the following arrangements:

```
(0), 1, 4, 5, 6, 7, 10, 11, 12, 15, 16, 19, (22)
(0), 1, 4, 5, 6, 7, 10, 12, 15, 16, 19, (22)
(0), 1, 4, 5, 7, 10, 11, 12, 15, 16, 19, (22)
(0), 1, 4, 5, 7, 10, 12, 15, 16, 19, (22)
(0), 1, 4, 6, 7, 10, 11, 12, 15, 16, 19, (22)
(0), 1, 4, 6, 7, 10, 12, 15, 16, 19, (22)
(0), 1, 4, 7, 10, 11, 12, 15, 16, 19, (22)
(0), 1, 4, 7, 10, 12, 15, 16, 19, (22)
```

(The charging outlet and your device's built-in adapter are shown in parentheses.) Given the adapters from the first example, the total number of arrangements that connect the charging outlet to your device is 8.

The second example above (the one that starts with 28, 33, 18) has many arrangements. Here are *a few*:

```
(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31, 32, 33, 34, 35, 38, 39, 42, 45, 46, 47, 48, 49, (52)
(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31, 32, 33, 34, 35, 38, 39, 42, 45, 46, 47, 49, (52)
(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31, 32, 33, 34, 35, 38, 39, 42, 45, 46, 48, 49, (52)
(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31, 32, 33, 34, 35, 38, 39, 42, 45, 46, 49, (52)
(0), 1, 2, 3, 4, 7, 8, 9, 10, 11, 14, 17, 18, 19, 20, 23, 24, 25, 28, 31, 32, 33, 34, 35, 38, 39, 42, 45, 47, 48, 49, (52)
(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45, 46, 48, 49, (52)
(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45, 46, 49, (52)
(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45, 47, 48, 49, (52)
(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45, 47, 49, (52)
(0), 3, 4, 7, 10, 11, 14, 17, 20, 23, 25, 28, 31, 34, 35, 38, 39, 42, 45, 48, 49, (52)
```

In total, this set of adapters can connect the charging outlet to your device in 19208 distinct arrangements.

You glance back down at your bag and try to remember why you brought so many adapters; there must be more than a trillion valid ways to arrange them! Surely, there must be an efficient way to count the arrangements.

What is the total number of distinct ways you can arrange the adapters to connect the charging outlet to your device?

#### Solution

We'll have to iterate recursively the adapters to find total ways to combine then. So the idea is, start from the beginning (from 0 to the whole sorted sequence of adapters until the device), step to step, find the next possible adapters and for each one, create a new path (the possible one and the rest of the adapters) do the same recursively, and count the distinct paths. Let's do it, with the cases above.

```{r}
library(tidyverse)

# this function chose, at next position (.i+1) in the adapters sequence
# which adapter can be used next (3 joltage range)
getNexts <- function(.dt, .i) {

  # indexes before the position of analysis
  .before <- 1:.i
  # what are the next adapters available in the joltage range (+3)
  .now <- (1:length(.dt))[which(.dt >= (.dt[.i]+1) & .dt <= (.dt[.i]+3))]

  # build maps of options
  indexes <- .now %>% 
    map(function(.n, .b, .s){
      c(.b,.n:.s)
    }, .b=.before, .s=length(.dt))

  # returns the adapters sequence
  indexes %>% 
    map(~.dt[.x]) %>% 
    return()
}

# this is a recursive function to iterate the adapters sequence
# to find all possible combinatinos
findAdapterComb <- function(.dt, .i=1){

  # if we are at the end of the sequence return NULL
  if(.i==length(.dt)) return(NULL)

  # find the combinations available at position .i
  comb.1 <- unique(getNexts(.dt, .i))

  # for each combination, find the combinations at position .i
  comb.2 <- comb.1 %>% 
    map(findAdapterComb, .i=.i+1) %>% 
    # removes the list levels and get unique values
    flatten() %>% 
    unique()

  # combine the options and returns
  c(comb.1, comb.2) %>% 
    unique() %>% 
    return()
}

# test case 01 = 8 possible answers

# reads the input as a vector of integers
input <- as.integer(readLines("./day10_test_01.txt"))

# adds the outlet and the device joltages and sort it
joltages <- sort(c(0, input, max(input)+3))

# counts the possible combinations
length(findAdapterComb(joltages))

# test case 02 = 19208 possible answers

# reads the input as a vector of integers
input <- as.integer(readLines("./day10_test_02.txt"))

# adds the outlet and the device joltages and sort it
joltages <- sort(c(0, input, max(input)+3))

# counts the possible combinations
length(findAdapterComb(joltages))
```

All is working, now let's do with the input data.

```{r}
# input data

# reads the input as a vector of integers
input <- as.integer(readLines("./day10_input.txt"))

# adds the outlet and the device joltages and sort it
joltages <- sort(c(0, input, max(input)+3))

# counts the possible combinations
# length(findAdapterComb(joltages)) >> YEAH, THIS DIDN'T WORK TOO MANY TIME TO PROCESS

```

Impossible!!! There is a more than 100 trillions of function call here, it will take ages, we need a better code.

In this case I consulted the masters and found the [puzzle answer from David Robinson](https://twitter.com/drob/status/1337103174868611074?s=20) that came with a more clean and compact code and a great optimization:

1. We doesn't return the adapters path along the calls, we need just count it (obviously). In this way save time returning data between the recursive calls
1. He came with the use of [`memoise package`] that transform a function in _memoised function_, a function that caches function calls so that if a previously seen set of inputs is seen, it can return the previously computed output.

First I didn't get why, in a recursive call sequence, this would be a vantage, but printing the parameters at start the call we can see that the same path of adapters is "calculated" over and over changing only the first adapter, so we can make some gains here, although takes a lot of time to process the input from this puzzles.

Let's see the David's code:

```{r}
library(memoise)
library(purrr)

# we creates a "cached version" of the function
countComb <- memoise(
  # this function return the # of possible combinations 
  # of adapters, it calculates the nexts possibles adapters from the start element
  # and calls itself from one of each.
  function(x){
    # if there is one element, returns there is *one* path
    if (length(x)==1) return(1)

    # which are the next possible adapters?
    alts <- which(x[2:length(x)]-x[1]<=3)

    # from each one call itself sum the answers and returns
    # tail do the trick to start from one of possible next adapters
    return(sum(map_dbl(alts, ~countComb(tail(x, -.))))) 
  })

# reads the input as a vector of integers
input <- as.integer(readLines("./day10_input.txt"))

# adds the outlet and the device joltages and sort it
joltages <- sort(c(0, input, max(input)+3))

resp <- countComb(joltages)
format(resp,scientific=FALSE)
```

This is amazing, is really, really fast! So, always consult [the masters](https://twitter.com/drob), you can learn a lot!

### 2020-12-22 AdvendOfCode2020/day_11.Rmd
Day 11 puzzle 1

---
title: "R Notebook"
output: html_notebook
---

--- Day 11: Seating System ---
Your plane lands with plenty of time to spare. The final leg of your journey is a ferry that goes directly to the tropical island where you can finally start your vacation. As you reach the waiting area to board the ferry, you realize you're so early, nobody else has even arrived yet!

By modeling the process people use to choose (or abandon) their seat in the waiting area, you're pretty sure you can predict the best place to sit. You make a quick map of the seat layout (your puzzle input).

The seat layout fits neatly on a grid. Each position is either floor (`.`), an empty seat (`L`), or an occupied seat (`#`). For example, the initial seat layout might look like this:

```
L.LL.LL.LL
LLLLLLL.LL
L.L.L..L..
LLLL.LL.LL
L.LL.LL.LL
L.LLLLL.LL
..L.L.....
LLLLLLLLLL
L.LLLLLL.L
L.LLLLL.LL
```

Now, you just need to model the people who will be arriving shortly. Fortunately, people are entirely predictable and always follow a simple set of rules. All decisions are based on the number of occupied seats adjacent to a given seat (one of the eight positions immediately up, down, left, right, or diagonal from the seat). The following rules are applied to every seat simultaneously:

```
If a seat is empty (`L`) and there are no occupied seats adjacent to it, the seat becomes occupied.
If a seat is occupied (`#`) and four or more seats adjacent to it are also occupied, the seat becomes empty.
Otherwise, the seat's state does not change.
Floor (`.`) never changes; seats don't move, and nobody sits on the floor.
```

After one round of these rules, every seat in the example layout becomes occupied:

```
#.##.##.##
#######.##
#.#.#..#..
####.##.##
#.##.##.##
#.#####.##
..#.#.....
##########
#.######.#
#.#####.##
```

After a second round, the seats with four or more occupied adjacent seats become empty again:

```
#.LL.L#.##
#LLLLLL.L#
L.L.L..L..
#LLL.LL.L#
#.LL.LL.LL
#.LLLL#.##
..L.L.....
#LLLLLLLL#
#.LLLLLL.L
#.#LLLL.##
```

This process continues for three more rounds:

```
#.##.L#.##
#L###LL.L#
L.#.#..#..
#L##.##.L#
#.##.LL.LL
#.###L#.##
..#.#.....
#L######L#
#.LL###L.L
#.#L###.##

#.#L.L#.##
#LLL#LL.L#
L.L.L..#..
#LLL.##.L#
#.LL.LL.LL
#.LL#L#.##
..L.L.....
#L#LLLL#L#
#.LLLLLL.L
#.#L#L#.##

#.#L.L#.##
#LLL#LL.L#
L.#.L..#..
#L##.##.L#
#.#L.LL.LL
#.#L#L#.##
..L.L.....
#L#L##L#L#
#.LLLLLL.L
#.#L#L#.##
```

At this point, something interesting happens: the chaos stabilizes and further applications of these rules cause no seats to change state! *Once people stop moving around, you count 37 occupied seats.*

Simulate your seating area by applying the seating rules repeatedly until no seats change state. How many seats end up occupied?

```{r message=FALSE, warning=FALSE}
# function that get the input data and converts
# into a int matrix of occupied seats
decodeInput <- function(input_txt){
  # slip each character as elements of a char vector
  char_seq <- unlist(strsplit(input_txt, ""))

  # converts the "char" array of ".", "L" and "#"
  # into a NA, 0 and 1 int values
  char_seq[char_seq=="L"] <- "0"
  char_seq[char_seq=="#"] <- "1"
  char_seq[char_seq=="."] <- "NA"

  # from array to a matrix
  return(matrix(as.integer(char_seq), nrow = length(input), byrow = T))
}

# calculates the sum of occupied seats adjacent to a seat position (x,y)
# and a seat map
sumAdjacentSeats <- function(x,y,seat_map){
  # get coordinates for adjacent seats
  dx <- x+(-1:1)
  dx <- dx[dx>=1 & dx<=nrow(seat_map)]  
  dy <- y+(-1:1)
  dy <- dy[dy>=1 & dy<=ncol(seat_map)]

  # the seat itself don't matter in the sum of adjacent
  seat_map[x,y] <- NA 
  # get the seats
  adj <- seat_map[dx,dy]
  # return the sum of occupied seats
  return(sum(adj,na.rm = T))
}

# given a seat (x,y) and a seat map returns
# an update for the seat following the rules 
updateSeatState <- function(x,y,seat_map){

  seat <- seat_map[x,y]

  # Floor (`.`/'NA') never changes; seats don't move, and nobody sits on the floor.
  if (is.na(seat)) return (NA)

  # Rule 01 - # If a seat is empty (`L`/'0') and
  # there are no occupied seats adjacent to it,
  # the seat becomes occupied.
  if (seat==0 & sumAdjacentSeats(x,y,seat_map)==0) return(1)

  # Rule 02 - If a seat is occupied (`#`/'1') and
  # four or more seats adjacent to it are also occupied, 
  # the seat becomes empty.
  if (seat==1 & sumAdjacentSeats(x,y,seat_map)>=4) return(0)

  # Otherwise, the seat's state does not change.
  return(seat)

}

# iterates until there is no update anymore in the seats map
iterateMap <- function(seat_map){
  # base map
  seatmap_update <- seat_map
  # empty map
  newSeatMap <- matrix(rep(1, prod(dim(seat_map))), nrow(seat_map))

  # while the map changes
  while(!all(newSeatMap==seatmap_update, na.rm = T)){

    # create a view of seat map to update
    newSeatMap <- seatmap_update

    # for each row..
    for (xi in 1:nrow(newSeatMap)) {
      # and each column...
      for (yi in 1:ncol(newSeatMap)) {
        # update the seat status
        seatmap_update[xi,yi] <- updateSeatState(xi, yi, newSeatMap)
      }
    }

  }

  return(seatmap_update)
}

# reads the input file as a vector of string
input <- readLines("./day11_test.txt")
# transform into a matrix
int_map <- decodeInput(input)
# iterate the seat occupation until stabilization
final_map <- iterateMap(int_map)
# Sums the occupied seats after map stabilization
sum(final_map, na.rm=T)

```
It's ok, so let's apply to the puzzle input.

```{r}
# reads the input file as a vector of string
input <- readLines("./day11_input.txt")
# transform into a matrix
int_map <- decodeInput(input)
# iterate the seat occupation until stabilization
final_map <- iterateMap(int_map)
# Sums the occupied seats after map stabilization
sum(final_map, na.rm=T)

```

### 2021-03-11 AdvendOfCode2020/day_11.Rmd
updates pendentes

@@
 
-```{r}
+```{r message=FALSE, warning=FALSE}
 # reads the input file as a vector of string
@@
 ```
+
+## Part Two
+
+--- Part Two ---
+As soon as people start to arrive, you realize your mistake. People don't just care about adjacent seats - they care about the first seat they can see in each of those eight directions!
+
+Now, instead of considering just the eight immediately adjacent seats, consider the first seat in each of those eight directions. For example, the empty seat below would see eight occupied seats:
+
+```
+.......#.
+...#.....
+.#.......
+.........
+..#L....#
+....#....
+.........
+#........
+...#.....
+```
+
+The leftmost empty seat below would only see one empty seat, but cannot see any of the occupied ones:
+
+```
+.............
+.L.L.#.#.#.#.
+.............
+```
+
+The empty seat below would see no occupied seats:
+
+```
+.##.##.
+#.#.#.#
+##...##
+...L...
+##...##
+#.#.#.#
+.##.##.
+```
+
+Also, people seem to be more tolerant than you expected: it now takes five or more visible occupied seats for an occupied seat to become empty (rather than four or more from the previous rules).
+The other rules still apply: empty seats that see no occupied seats become occupied, seats matching no rule don't change, and floor never changes.
+
+Given the same starting layout as above, these new rules cause the seating area to shift around as follows:
+
+```
+L.LL.LL.LL
+LLLLLLL.LL
+L.L.L..L..
+LLLL.LL.LL
+L.LL.LL.LL
+L.LLLLL.LL
+..L.L.....
+LLLLLLLLLL
+L.LLLLLL.L
+L.LLLLL.LL
+```
+
+```
+#.##.##.##
+#######.##
+#.#.#..#..
+####.##.##
+#.##.##.##
+#.#####.##
+..#.#.....
+##########
+#.######.#
+#.#####.##
+```
+
+```
+#.LL.LL.L#
+#LLLLLL.LL
+L.L.L..L..
+LLLL.LL.LL
+L.LL.LL.LL
+L.LLLLL.LL
+..L.L.....
+LLLLLLLLL#
+#.LLLLLL.L
+#.LLLLL.L#
+```
+
+```
+#.L#.##.L#
+#L#####.LL
+L.#.#..#..
+##L#.##.##
+#.##.#L.##
+#.#####.#L
+..#.#.....
+LLL####LL#
+#.L#####.L
+#.L####.L#
+```
+
+```
+#.L#.L#.L#
+#LLLLLL.LL
+L.L.L..#..
+##LL.LL.L#
+L.LL.LL.L#
+#.LLLLL.LL
+..L.L.....
+LLLLLLLLL#
+#.LLLLL#.L
+#.L#LL#.L#
+```
+
+```
+#.L#.L#.L#
+#LLLLLL.LL
+L.L.L..#..
+##L#.#L.L#
+L.L#.#L.L#
+#.L####.LL
+..#.#.....
+LLL###LLL#
+#.LLLLL#.L
+#.L#LL#.L#
+
+```
+
+```
+#.L#.L#.L#
+#LLLLLL.LL
+L.L.L..#..
+##L#.#L.L#
+L.L#.LL.L#
+#.LLLL#.LL
+..#.L.....
+LLL###LLL#
+#.LLLLL#.L
+#.L#LL#.L#
+```
+
+Again, at this point, people stop shifting around and the seating area reaches equilibrium. Once this occurs, you count 26 occupied seats.
+
+Given the new visibility method and the rule change for occupied seats becoming empty, once equilibrium is reached, how many seats end up occupied?

### 2021-03-11 ele_ela_gutenberg/book_text_import.R
scripts de dowload e importacao de livros do projeto gutenberg

@@ -1,0 +1,51 @@
+library(tidyverse)
+
+books <- readRDS("./ele_ela_gutenberg/books.rds")
+book_files <- readRDS("./ele_ela_gutenberg/books_file.rds")
+
+bks <- books %>% 
+  select(book_id, author, title, subject, language) %>% 
+  unnest(c(author, title, subject, language)) %>% 
+  distinct() %>% 
+  filter( str_detect(str_to_lower(subject), "fiction"),
+          language == "Portuguese" ) %>% 
+  inner_join(book_files, by="book_id")
+
+book_lines <- bks %>% 
+  split(1:nrow(.)) %>% 
+  map_df(function(.book){
+
+    raw_text <- readLines(.book$filename, encoding = "UTF-8")
+    
+    text_lines <- raw_text %>% 
+      str_c(" ") %>% 
+      str_replace("^ $","<empty_line>") %>% 
+      str_c(collapse = "") %>% 
+      str_split("<empty_line>") %>% 
+      unlist() %>% 
+      str_trim()
+    
+    text_lines <- text_lines[text_lines!=""]
+    
+    tibble(line_text = text_lines) %>% 
+      mutate(
+        book_id = .book$book_id,
+        lines = length(text_lines),
+        line_number = row_number()
+      ) %>% 
+      nest(text_lines=c(line_number, line_text)) %>% 
+      return()
+    
+    # .book %>%
+    #   select(book_id) %>% 
+    #   mutate( text = list(text_lines),
+    #           lines = length(text_lines)) %>% 
+    #   unnest(text) %>% 
+    #   mutate( line_number = row_number() ) %>% 
+    #   nest(text_lines=c(line_number, text)) %>% 
+    #   return()
+    
+  })
+
+book_lines %>% 
+  saveRDS("./ele_ela_gutenberg/fiction_book_text.rds")

### 2021-03-11 ele_ela_gutenberg/index_book_download.R

@@ -1,0 +1,104 @@
+library(tidyverse)
+library(glue)
+library(rvest)
+library(janitor)
+
+# portuguese directory
+url <- "https://www.gutenberg.org/browse/languages/pt"
+index_page <- read_html(url)
+
+# extracting links to books
+book_links <- index_page %>% 
+    html_nodes("body > div.container > div > div.pgdbbylanguage > ul > li > a") %>%
+    as.character()
+
+# creating a book dataset
+books <- tibble(
+    book_link = book_links
+  ) %>% 
+  mutate(
+    # extract book number
+    book_id = str_remove_all(str_extract(book_links,"\".+\""),"[^\\d]"),
+    # extract book title
+    book_title = str_remove_all(str_extract(book_links,">.+<"),"[<>]")
+  ) %>% 
+  relocate(book_id, book_title)
+
+
+# for each book get the metadata info
+books_info_values <- books$book_id %>% 
+  map_df(function(.bid){
+    
+    # book download page
+    meta_url <- glue("https://www.gutenberg.org/ebooks/{.bid}")
+    meta_page <- read_html(meta_url)
+    
+    # there is two tables
+    # #1 download liks 
+    # #2 book info
+    page_tables <- html_table(meta_page)
+    
+    # download links
+    download_links <- page_tables[[1]] %>% 
+      clean_names() %>% 
+      select(format, url, size) %>% 
+      mutate( format = make_clean_names(format) )
+    
+    text_link <- download_links %>% 
+      filter(format=="plain_text_utf_8") %>% 
+      pull(url)
+    
+    # extract info table
+    book_info <-  page_tables[[2]] %>% 
+      as_tibble() %>% 
+      # it is a key-value table
+      set_names(c("names","values")) %>% 
+      # add meta_url
+      bind_rows(
+        tibble(
+          names = c("book_page","download_links"),
+          values = c(meta_url,text_link)
+        )
+      ) %>%  
+      mutate(
+        # clean key names
+        names  = make_clean_names(names),
+        # remove \r from the strings values 
+        values = str_replace_all(values, "\\r"," "),
+        # add book_id as register key
+        book_id = as.integer(.bid)
+      ) %>% 
+      relocate(book_id)
+
+    return(book_info)
+    
+  })
+
+books_df <- books_info_values %>% 
[…]

### 2021-11-16 2020_09_22_tidy_tuesday.Rmd
final backup

---
title: "TidyTemplate"
date: 2020-09-24
output: html_output
editor_options: 
  chunk_output_type: console
---

# TidyTuesday

Join the R4DS Online Learning Community in the weekly #TidyTuesday event!
Every week we post a raw dataset, a chart or article related to that dataset, and ask you to explore the data.
While the dataset will be “tamed”, it will not always be tidy! As such you might need to apply various R for Data Science techniques to wrangle the data into a true tidy format.
The goal of TidyTuesday is to apply your R skills, get feedback, explore other’s work, and connect with the greater #RStats community!
As such we encourage everyone of all skills to participate!

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(tidytuesdayR)
library(scales)
theme_set(theme_light())

```

# Load the weekly Data

Dowload the weekly data and make available in the `tt` object.

```{r Load}

tt <- tt_load("2020-09-22")

tt

```

# Readme

Take a look at the readme for the weekly data to get insight on the dataset.
This includes a data dictionary, source, and a link to an article on the data.

```{r Readme, eval = interactive()}

peaks <- tt$peaks

peaks %>% 
  arrange(desc(height_metres)) %>% 
  head(30) %>% 
  mutate(peak_name = fct_reorder(peak_name, height_metres)) %>% 
  ggplot(aes(height_metres, peak_name, fill= climbing_status)) +
  geom_col() +
  labs(x="Height (meters)",
       y="",
       title="Tallest peaks in the Himalayas",
       fill="")

```

### 2021-11-16 animal_crossing_portfolio/how_i_made_my_millions.Rmd
final backup

---
title: "R Notebook"
output: html_notebook
---

Reprodução do post [#TidyTuesday - Building Stalk Portfolios with R](https://www.robert-hickman.eu/post/how_i_made_my_millions/)

In Animal Crossing, one aspect in particular that captured my attention was the weekly [market for turnips](https://animalcrossing.fandom.com/wiki/White_turnip) on each player’s island. Every Sunday, the player has the opportunity to buy turnips from a salesperson, which they then have a week to sell (before the turnips rot). The prices of turnips fluctuate over the week and (as far as I know) the vegetables have little function outside of buying/selling, so in essence they work as stocks which can be bought once per week, and the player must clear before the next offering.

* players can buy as many turnips as they want from character a for price x per turnip* on Sunday morning
* players can then sell as many or as few turnips as they want to character b for price y(t) from the morning until 10pm
* prices vary depending on price y(t-1) and a given pattern, changing at the start of each day and then at midday each day (so there are 12 independent selling prices y through a week)
* on the next Sunday everything resets
* for the next week the pattern the prices follow may or may not be different

Modelling the movement of the turnip price leads to some interesting analysis using models from financial data science, and while this data wasn’t a part of #TidyTuesday, I think it’s related enough to potentially be of interest to people.

```{r setup}
#load libraries
library(tidyverse) 
library(MASS)
library(conflicted)
#prefer tidyverse functions
preferred <- map(c("filter", "select", "lag"), conflict_prefer, "dplyr")

#source the functions governing turnips price dynamics
source("./turnip_funs.R")
```

After purchasing turnips the prices can either:

* fluctuate (go up and down around mean $100 purchase price)
* ‘spike’ upwards around midweek (where prices will decrease before shooting up past the ~$100 purchase price around Wednesday). This is actually two separate but similar patterns, where ‘large spike’ leads to greater selling prices than ‘small spike’
* decreasing in which the player has no hope of making a profit that week as prices will decrease every day before resetting on the Sunday

The patterns are not completely memory-less, and progress as a Markov Chain, where the probability of prices following a pattern next week are dependent on the current price pattern.

```{r}

#the four patterns
states <- c("fluctuating", "large_spike", "decreasing", "small_spike")

#build the transition matrix between the states
transition_matrix <- matrix(
  c(
    0.2, 0.3, 0.15, 0.35,
    0.5, 0.05, 0.2, 0.25,
    0.25, 0.45, 0.05, 0.25,
    0.45, 0.25, 0.15, 0.15
  ),
  nrow = 4, byrow = TRUE)

#name the current (rows) and next (cols) states
rownames(transition_matrix) <- states
colnames(transition_matrix) <- states

transition_matrix
```

If we are a naive observer, the chance of observing any pattern is therefore solved by taking the left eigenvectors of this matrix:

```{r}
#take the elft eignevector
#ginv from the MASS package
left_eigen <- ginv(eigen(transition_matrix)$vectors)[1,]
pattern_likelihood_analytic <- left_eigen / sum(left_eigen)
#name the probabilities
names(pattern_likelihood_analytic) <- states

pattern_likelihood_analytic
```

Where we see that around half the time we have a chance of either a large or a small spike in prices around midweek (24.7% + 25.9%), with the majority of the remaining weeks showing a fluctuating pattern (where the player can still make a small profit). The worst case scenario of continually decreasing prices happens only 14.7% of the time, so overall, the stalk market looks like a pretty good bet for investors.

Of course, we can also do this using Hamiltonian Monte Carlo methods by simulating a few sets of independent weeks

```{r warning=FALSE, message=FALSE}
#transition probabilities
transition_df <- as.data.frame(transition_matrix) %>%
  rownames_to_column(var = "current_state") %>%
  pivot_longer(cols = all_of(states), names_to = "next_state", values_to = "prob") %>%
  group_by(current_state) %>%
  mutate(cum_prob = cumsum(prob)) %>%
  ungroup()

#get the next pattern from the current pattern
find_next_pattern <- function(pattern, rng, transitions = transition_df) {
  next_transition <- transitions %>%
    #find possible patterns
    filter(current_state == pattern & cum_prob > rng) %>%
    #take top row
    .[1,]
  #next state is that pattern
  next_state <- next_transition$next_state
}

#run forward for prop_forward weeks for each run to check convergence
transition_patterns <- function(initial_pattern, prop_forward) {
  patterns <- c()
  pattern <- initial_pattern
  #run n times
  for(runs in seq(prop_forward)) {
    pattern <- find_next_pattern(pattern, runif(1))
    patterns <- append(patterns, pattern)
  }
  #return as df
  df <- data.frame(
    initial_pattern,
    pattern = as.character(patterns),
    t = 1:prop_forward
  )
  return(df)
}

#repeat sims n times
simulation_reps <- 100
#how many weeks to run each sim for
prop_forward = 10
#run the sims
pattern_likelihood <- states %>%
  rep(simulation_reps) %>%
  map_df(., transition_patterns, prop_forward) %>%
  group_by(pattern) %>%
  summarise(prob = n() / (simulation_reps *  prop_forward * length(states)))

pattern_likelihood

```

And we get pretty much the same numbers (as we would expect). To show the relative frequencies and how well our two methods of finding the stationary probabilities work, we can easily graph this using ggplot

```{r}
p1 <- pattern_likelihood_analytic %>%
  as.data.frame() %>%
  rownames_to_column("pattern") %>%
  left_join(pattern_likelihood, by = "pattern") %>%
  rename(hmc = "prob", analytic = ".") %>%
  pivot_longer(c("hmc", "analytic"), names_to = "calc", values_to = "prob") %>%
  ggplot(aes(x = pattern, y = prob, group = calc)) +
  geom_bar(stat = "identity", position = "dodge", aes(fill = calc), colour = "black") +
  scale_fill_manual(values = c("dodgerblue", "orange")) +
  labs(
    title = "probability of observing any one price pattern when randomly sampling",
    subtitle = "showing difference estimate from analytic and Monte Carlo methods",
    x = "week's prices pattern",
    y = "probability"
  ) +
  theme_minimal()

p1
```

So, given the likelihood of spikes in prices, we know we’ve got a good chance of making some money by buying and selling turnips.

# Modelling Turnips Prices

To calculate exactly how much we might expect, it’s easiest, just to simulate the prices a load of times

The function randomly selects an initial (Sunday) price for turnips to be bought at, and then, runs the simulation code for a given pattern of prices. The second argument simply gives a list of names for the epochs (each day for both AM or PM, which will have different selling prices). I wrap the simulation up into a df because I find it easier to work with though the real meat of the simulation is a vector of length 14 which contains the ‘two’ Sunday buying prices (which will be identical- it’s just to make it easier for me to count), and the 12 selling prices from Monday AM - Saturday PM.

```{r}
#the epochs for buying and selling turnips
#14 epochs, 2 identical buying epochs, and 12 unique selling epochs
week <- c("sun", "mon", "tues", "wed", "thurs", "fri", "sat")
epochs <- paste(rep(week, each = 2), rep(c("AM", "PM"), 7))

#simulate a week of prices given a pattern
simulate_week <- function(pattern, epochs) {
  #set up prices vector
  sunday_price <- sample(90:110, 1)
  initial_prices <- c(rep(sunday_price, 2), rep(0, 12))

  #simulate pattern
  if(pattern == "decreasing") {
    week_prices <- sim_decreasing(
      prices = initial_prices
    )
  } else if(pattern == "fluctuating") {
    week_prices <- sim_fluctuating(
      prices = initial_prices, 
      first_epochs = c(sample(0:6, 1), sample(2:3, 1))
    )
  } else if(pattern == "large_spike") {
    week_prices <- sim_largespike(
      prices = initial_prices, 
      rate = runif(1, 0.85, 0.95), 
      first_peak = sample(2:8, 1)
    )
  } else if(pattern == "small_spike") {
    week_prices <- sim_smallspike(
      prices = initial_prices, 
      first_peak = sample(1:8, 1)
    )
  }

  #arrange df
  weekly_prices <- data.frame(
    day = epochs,
    buy_price = sunday_price,
    price = week_prices
  )

  return(weekly_prices)
}
```

We can calculate how many times each pattern should be run by defining the number of simulations we want to run, and sampling price patterns, weighted by likelihood, from the df we calculated above.

Then we just have to sample the vector of 1000 choices of the 4 patterns to the function and do a little munging at the end. After we can get a sense of which days are most profitable for selling turnips by plotting the histogram of the return (by how many times we have multiplied our original stock of $) on turnip investment.

```{r warning=FALSE, message=FALSE}
#how many simulations of weeks to run
simulation_reps <- 1000
prices <- pattern_likelihood %>%
  #sample patterns by likelihood
  sample_n(simulation_reps, weight = prob, replace = TRUE) %>%
  .$pattern %>%
  map_df(., simulate_week, epochs) %>%
  mutate(return = price / buy_price,
         day = factor(day, levels = epochs)) %>%
  filter(!grepl("sun [A-Z]{2}", day)) %>%
  group_by(day) %>%
  mutate(mean_return = mean(return)) %>%
  ungroup()

p2 <- ggplot(prices, aes(x = return, fill = mean_return)) +
  geom_histogram(alpha = 0.8, colour = "black") +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "dodgerblue", size = 1) +
  scale_fill_gradient2(low = "red", high = "green", mid = "black", midpoint = 1) +
  scale_x_continuous(limits = c(0, 2)) +
  theme_minimal() +
  labs(
    title = "which days yield greatest profits in the stalk market?",
    subtitle = paste("based on", simulation_reps, "simulations"),
    x = "return on turnip investment",
    y = "observed count"
  ) +
  facet_wrap(~day)

p2
```

he histograms struggle to portray the movement of time across epochs, so I also wanted to plot the prices using the ggridges package to produce density joy plots over time. Here we can see a bit clearer that it’s only the long positive tails on the distributions which give us an expected return slightly above break even from Wednesday AM- Thursday PM:

```{r warning=FALSE, message=FALSE}
library(ggridges)

p3 <- ggplot(prices, aes(x = return, y = day, fill = mean_return)) +
  geom_density_ridges2() +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "dodgerblue", size = 1) +
  scale_fill_gradient2(low = "red", high = "green", mid = "black", midpoint = 1) +
  scale_x_continuous(limits = c(0, 2.5)) +
  labs(
    title = "which days yield greatest profits in the stalk market?",
    subtitle = paste("based on", simulation_reps, "simulations"),
    x = "return on turnip investment",
    y = "observed density by day"
  ) +
  theme_minimal()

p3
```

In addition to performing manual tasks or investing in turnip stocks, the player can also bank their hard-earned money and collect the interest. As far as I can tell banked money earns interest at a rate of 0.05%* which is payed out monthly.

To work with this a bit easier, I’m going to make a slight tweak and calculate as if the interested was earned daily. Therefore, if we take time 0 to be a Sunday morning at the beginning of the month, by the following Saturday, the player who put $100 dollars in the bank will know have (1 + (0.05/ 100)) ^ (6/30) * 100 in their bank account (or an extra 1cent if you calculate).

The risk-free return over that one week will therefore have been 1cent- it is the return the player receives without having to risk their money buying/selling turnips. For each day over the week, this risk-free return is easy to calculate. Because we are only interested in this compared to the returns on investing savings in turnips, the amount in the bank doesn’t actually matter- we only care on the interest gained as a proportion of savings.

*the FT has an article on the recent Animal Crossing interest rate cut and why it forces players into riskier assets like turnips

```{r}

monthly_interest <- 1.005

interest_df <- data.frame(day = factor(epochs, levels = epochs)) %>%
  mutate(interest_days = rep(0:6, each = 2)) %>%
  mutate(interest_gained = (1 * (monthly_interest ^ (1/30)) ^ interest_days) -1)

interest_df

```

We can then work out the return on investment compared to risk. The simplest way to do this is to use the *Sharpe ratio* which can be formalized as:

$S_a={E[R_a-R_b]}/{\sqrt{var[R_a-R_b]}}$

Where we calculate the Sharpe ratio S of an asset a which is a function of the expected excess return (aka profit) R of that asset above the expected excess return of a ‘safe’ asset b (in this case the interest on money in the bank).

This is then divided by the variance of the expected gain above the risk free asset. We call the difference in return of the risky and safe asset Ra - Rb the ‘excess return’.

It should also be clear that we want a Sharpe ratio of at least greater than 0 to make our investment worthwhile (as a risky asset is time discounted and the possibility of prospect-like losses); generally we want a Sharpe ratio of 1 to indicate a good investment.

As we have the interest gained per day, we can calculate the excess return by joining our interest df and taking the return by day for each simulation as Ra, which we average to find the expected and variance.

```{r}
#join in interest data
Sharpe_mean_returns <- prices %>%
  left_join(interest_df, by = "day") %>%
  #calculate excess return over safe asset
  mutate(excess_return = (return - 1) - interest_gained) %>%
  group_by(day) %>%
  #calc nominator and denominator
  summarise(mean_excess = mean(excess_return),
            sd_excess = sd(excess_return)) %>%
  mutate(sharpe_ratio = mean_excess / sd_excess)

select(Sharpe_mean_returns, day, sharpe_ratio)
```

So, as expected, the only epochs which show a positive Sharpe ratio are in the middle of the week, where prices spike. If we plot this we get a clearer indication of this:

```{r}
p4 <- ggplot(Sharpe_mean_returns, aes(x = day, y = sharpe_ratio, group = 1)) +
  geom_line(colour = "dodgerblue") +
  geom_point(size = 2, colour = "dodgerblue") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "orange") +
  labs(
    title = "Sharpe ratio for selling turnip investment on a given epoch",
    subtitle = paste("based on", simulation_reps, "simulations"),
    x = "day",
    y = "Sharpe ratio"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p4
```

Remember, we want a Sharpe ratio of around 1 to be confident our investment is a good idea, so for the casual investor, the stalk market doesn’t seem to give good value. If a player did want to dabble and could only dedicate so many hours per week to the game, the calculation suggests they should try to make time somewhere around Wednesday or Thursday to check in on the market to maximize their returns.

# Globalizing the Stalk Market

However, all is not lost for turnip investors! While each player lives and works on their own island, they are able to visit the islands of their friends and can buy and sell Turnips there following the same rules. However, prices (both buying on Sunday and selling for the rest of the week) are independent between player’s islands, so a smart investor can sample many markets and sell in the one which offers the greatest price.

To simulate these open markets, we image a player has n friends who will let them visit, and calculate the potential profit made by an optimal investor (buying on the cheapest island, and selling on the most profitable). We simulate each epoch on each island the same 1000 times and take the max potential returns per epoch.

```{r}
#do the same for n islands
simulate_open_economies <- function(islands, pattern_likelihood, epochs) {
  patterns <- pattern_likelihood %>%
    sample_n(islands, weight = prob, replace = TRUE) %>%
    .$pattern

  #munge
  all_prices <- map_df(patterns, simulate_week, epochs) %>%
    #will always buy at lowest and sell at highest
    mutate(buy_price = min(buy_price)) %>%
    group_by(day) %>%
    mutate(sell_price = max(price)) %>%
    select(day, buy_price, sell_price) %>%
    unique() %>%
    mutate(islands)
}

#run on 1:n islands
n_islands <- 10
open_prices <- rep(1:n_islands, simulation_reps) %>%
  map_df(simulate_open_economies, pattern_likelihood, epochs) %>%
  group_by(islands) %>%
  mutate(return = sell_price / buy_price,
         day = factor(day, levels = epochs)) %>%
  filter(!grepl("sun [A-Z]{2}", day)) %>%
  group_by(islands, day) %>%
  mutate(mean_return = mean(return)) %>%
  ungroup()
```

If we then paste the density of returns by simulation, colored by the number of islands, we can see a clear rightward shift towards greater returns when a player has more friends. To think about this simply, if we imagine a player has infinite friends, they will always buy turnips for the minimum possible price ($90) and always sell them for the maximum possible price on that day.

```{r warning=FALSE, message=FALSE}
p5 <- ggplot(open_prices, aes(x = return, y = day, group = paste(day, islands),
                              fill = islands, colour = islands)) +
  geom_density_ridges2(alpha = 0.2) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "dodgerblue", size = 1) +
  scale_fill_gradient(low = "yellow", high = "green") +
  scale_colour_gradient(low = "yellow", high = "green") +
  scale_x_continuous(limits = c(0, 2.5)) +
  labs(
    title = "which days yield greatest profits in the stalk market?",
    subtitle = "by day and number of islands sampled",
    x = "return on investment",
    y = "day"
  ) +
  theme_minimal()

p5
```

What’s striking is that even with just 10 friends, player can be pretty much guaranteed to always make profit no matter which day they collude to all check their islands prices- the mean return on investment is clearly above 1.0 even by the following Saturday. The best potential returns are clearly still to be had midweek however, where now a player can clearly expect a doubling of their investment:

```{r}
open_prices %>%
  group_by(islands, day) %>%
  summarise(mean_return = mean(return)) %>%
  arrange(-mean_return) %>%
  head(n = 10)
```

If we use our Sharpe ratio calculation to then calculate when a player should collude with friends to all check their local turnip prices* we might expect therefore that it will also suggest checking somewhere in this midweek spike. However, if we plot it, we find an unexpected result:

*if we assume that quarantined players probably can manage to check prices more than once/twice a week, managing to co-ordinate between multiple players is going to get very hard very quickly so this constraint really will become a factor

```{r}
#calculate Sharpe ratio per island as before
Sharpe_mean_open_returns <- open_prices %>%
  left_join(interest_df, by = "day") %>%
  mutate(excess_return = (return - 1) - interest_gained) %>%
  group_by(islands, day) %>%
  summarise(mean_excess = mean(excess_return),
            sd_excess = sd(excess_return)) %>%
  mutate(sharpe_ratio = mean_excess / sd_excess)

#plot the sharpe ratio coloured by islands
p6 <- ggplot(Sharpe_mean_open_returns, 
             aes(x = day, y = sharpe_ratio, group = islands, colour = islands)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "orange") +
  labs(
    title = "Sharpe ratio for selling turnip investment on a given epoch",
    subtitle = "by day and number of islands sampled",
    x = "day",
    y = "Sharpe ratio"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p6
```

As we increase the number of islands, the Sharpe ratio tilts towards selling earlier (when the mean returns are lower). Why do we get this result? To investigate we need to look at some real stock charts, which we can easily do via the tidyquant package.

```{r message=FALSE, warning=FALSE}
library(tidyquant)

#get nintendo stock data
nintendo_symbol <- "7974.T"
nintendo <- getSymbols(nintendo_symbol,
                       from = '2019-06-01',
                       to = "2020-03-15",
                       auto.assign = FALSE) %>%
  as.data.frame() %>%
  rename_all(~gsub(paste0(nintendo_symbol, "."), "", .)) %>%
  rownames_to_column("date") %>%
  mutate(date = as.Date(date)) %>%
  select(date, close = Close)

#plot the last 9 months of nintendo stock
p7 <- ggplot(nintendo, aes(x = date, y = close)) +
  geom_line(size = 2, colour = "orange") +
  labs(
    title = "Nintendo stock prices",
    subtitle = "data from June 2019-March 2020",
    x = "date",
    y = "closing stock price"
  ) +
  theme_minimal()

p7
```

Even though the Nintendo stock price has been fairly volatile over the past 9 months, it’s held its value fairly steadily- it would be literally impossible for a trader to double their investment over this time (c.f. turnips in a week). We can drill down into the data by looking at the daily change in closing price (as a fraction of the price), using dplyr::lag()

```{r message=FALSE, warning=FALSE}
p8 <- nintendo %>%
  mutate(daily_change = (close - lag(close)) / lag(close)) %>%
  ggplot(aes(x = daily_change)) +
  geom_histogram(fill = "orange", colour = "black") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "dodgerblue") +
  labs(
    title = "daily changes in Nintendo stock prices",
    subtitle = "data from June 2019-March 2020",
    x = "fractional daily change in price",
    y = "count"
  ) +
  theme_minimal()

p8
```

There’s three things to note here: - the daily change in prices (roughly) follows a normal distribution* (with a mean slightly above 0 over a given time frame) - the daily change in price is fairly small, i.e. the price is fairly stable - there is a greater downside risk in large price moves- you’re more likely to see a big reduction in price than a big increase (see the recent downturn in Nintendo stock due to COVID-19 for example**)

*stock returns don’t follow a normal distribution if you rigorously test it, but it’s close enough to be useful

**yes, I know it has bounced back up, the data was selectively chosen to make a point

These are important basic heuristics for portfolio building and we can see that our stalk market fails on all three. Luckily, the fact that our turnip returns are skewed (even if they are skewed upwards instead of downwards as in the real life data).

Instead of using the Sharpe ratio, which considers the total standard deviation of the returns, we can use the Sortino ratio which is a risk-adjusted version to control for the downside risk of investment (i.e. you’re more likely to make big losses than big gains). We know that with multiple friends, we can be pretty confident of making big returns,

$S_a={E(R_a-MAR)}/{\sqrt{{1/n}*\int_{-\infty}^{MAR}{(MAR-R_a)^2dR}}}$

which ok, looks pretty rough, but is simple enough to calculate.

The numerator is just the same as the Sharpe ratio numerator, except instead of the returns on asset a vs. a risk-free asset, we’re now calculating the returns vs. a Minimal Acceptable Return (MAR). Re-imagine our scenario where someone only has x hours spare to play Animal Crossing, they aren’t going to go through the stress and commitment to play the stalk market without making at least MAR returns (where MAR is some number).

The denominator also looks more complicated than the Sharpe ratio, but remember, for that we want to find

$demon_{Sharp}=\sqrt{var[R_a-R_b]}=sd[R_a-R_b]$

which is what we’re calculating here, just we are limiting the standard deviation to the downside risk, which means we only take the standard deviation of returns which fall beneath the MAR (hence the max argument in the integration).

For instance, let’s say we want an excess return of 1, i.e. we want to judge the profitability of checking certain epochs to at least double our initial investment on turnips:

```{r}
#want to double investment so MAR = 1
MAR <- 1

#calc Sortino ratio
Sortino_ratio <- open_prices %>%
  group_by(day, islands) %>%
  mutate(excess_return = return - 1) %>%
  summarise(
    #numerator
    mean_excess = mean(excess_return - MAR),
    #denominator squared for readability
    downside_sq = sum((MAR - excess_return[excess_return < MAR])^2/n())
  ) %>%
  #calc
  mutate(sortino_ratio = mean_excess / sqrt(downside_sq))

#plot the Sortino ratio by epoch
p9 <- ggplot(Sortino_ratio, aes(x = day, y = sortino_ratio, group = islands, colour = islands)) +
  geom_line() +
  geom_point(size = 2) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "orange") +
  labs(
    title = "Sortino ratio for selling turnip investment on a given epoch",
    subtitle = "by day and number of islands sampled",
    x = "day",
    y = "Sortino ratio"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p9
```

### 2021-11-16 GAM/motorcycle_crash.R

@@ -1,0 +1,55 @@
+# database
+mcycle <- MASS::mcycle
+head(mcycle)
+plot(mcycle$times, mcycle$accel)
+
+
+# linear model
+lm_mod <- lm(accel~times, data=mcycle)
+termplot(lm_mod, partial.resid = T, se=T)
+
+# gam model
+library(mgcv)
+gam_mod <- gam(accel ~ s(times), data=mcycle )
+plot(gam_mod, residuals = T, pch=1)
+summary(gam_mod)
+
+# base functions
+coef(gam_mod)
+
+# smoothing parameter
+# FIT = LIKEHOOD - (LAMBDA)*WIGGLINESS
+gm_auto <- gam(accel ~ s(times), data = mcycle)
+gm_sp1  <- gam(accel ~ s(times, sp = 0.42), data = mcycle)
+gm_reml <- gam(accel ~ s(times), data = mcycle, method = "REML")
+
+par(mfrow = c(1, 3))
+plot(gm_auto, residuals = TRUE, pch = 1)
+plot(gm_sp1,  residuals = TRUE, pch = 1)
+plot(gm_reml, residuals = TRUE, pch = 1)
+
+str(gm_sp1)
+
+
+# number of basis function
+
+# Fit a GAM with 3 basis functions
+gam_mod_k3 <- gam(accel ~ s(times, k = 3), data = mcycle)
+
+# Fit with 20 basis functions
+gam_mod_k20 <- gam(accel ~ s(times, k = 20), data = mcycle)
+
+# Visualize the GAMs
+par(mfrow = c(1, 2))
+plot(gam_mod_k3, residuals = TRUE, pch = 1)
+plot(gam_mod_k20, residuals = TRUE, pch = 1)
+
+# visualizing smooth parameter
+gm_reml$smooth[[1]]$sp
+
+
+# Fit the GAM with 50 base
+gam_mod_sk <- gam(accel~s(times, k=50), data=mcycle, sp=0.0001)
+
+# Visualize the model
+plot(gam_mod_sk, residuals = TRUE, pch = 1)

### 2021-11-16 GAM/mpg.R

@@ -1,0 +1,69 @@
+library(mgcv)
+#library(gamair)
+data("mpg", package="gamair")
+
+head(mpg)
+plot(mpg$weight, mpg$hw.mpg)
+
+# one variable
+model <- gam(hw.mpg ~ s(weight), data=mpg, method="REML")
+par(mfrow = c(1, 1))
+plot(model, residuals = T, pch=1)
+
+# two variable
+model2 <- gam(hw.mpg ~ s(weight) + s(length), data=mpg, method="REML")
+par(mfrow = c(1, 2))
+plot(model2, residuals = T, pch=1)
+
+# two variable (one linear)
+model3 <- gam(hw.mpg ~ length + s(weight), data=mpg, method="REML")
+par(mfrow = c(1, 1))
+plot(model3, residuals = T, pch=1)
+
+# categorical variables
+str(mpg)
+
+model3 <- gam(hw.mpg ~ s(weight) + fuel, data = mpg, method = "REML")
+par(mfrow = c(1, 2))
+plot(model3, residuals = T, pch=1)
+termplot(model3)
+
+model4 <- gam(hw.mpg ~ s(weight, by = fuel), data = mpg, method = "REML")
+plot(model4, residuals = T, pch=1)
+termplot(model3, se = T)
+
+model4b <- gam(hw.mpg ~ s(weight, by = fuel) + fuel, data = mpg, method = "REML")
+par(mfrow = c(1, 3))
+plot(model4b, residuals = T, pch=1)
+termplot(model4b, se = T)
+
+## City
+
+# Examine the data
+head(mpg)
+str(mpg)
+
+# Fit the model
+mod_city <- gam(city.mpg ~ s(weight) + s(length) + s(price), 
+                data = mpg, method = "REML")
+
+# Plot the model
+#par(mfrow = c(2, 2))
+plot(mod_city, pages = 1)
+
+
+# Fit the model
+mod_city3 <- gam(city.mpg ~ s(weight, by=drive) + s(length, by=drive) + s(price, by=drive) + drive, 
+                 data = mpg, method = "REML")
+
+# Plot the model
+plot(mod_city3, pages = 1)
+
+
+### Interpreting
+
+mod_hwy <- gam(hw.mpg ~ s(weight) + s(rpm) +  s(price) + s(comp.ratio) + s(width) + fuel + cylinders,
+               data = mpg, method = "REML")
+
+summary(mod_hwy)
+plot(mod_hwy, pages=1)

### 2021-11-16 ele_ela_gutenberg/book_analysis.R

@@ -1,0 +1,46 @@
+library(tidyverse)
+library(tidytext)
+
+books <- readRDS("./ele_ela_gutenberg/book_text.rds")
+
+bigrams <- books %>% 
+  unnest(text_lines) %>% 
+  mutate(line_text = str_to_lower(line_text)) %>% 
+  unnest_tokens(bigram, line_text, token = "ngrams", n=2, collapse = F)
+
+bigrams_separated <- bigrams %>% 
+  separate(bigram, c("word1","word2"), sep=" ")
+
+
+saveRDS(bigrams_separated,"./ele_ela_gutenberg/book_bigrams.rds")
+
+he_she_words <- bigrams_separated %>% 
+  filter( word1 %in% c("ele","ela", "eles","elas") ) %>% 
+  mutate( word1 = str_remove(word1,"s") )
+
+he_she_words %>% 
+  filter(!word2 %in% c(tm::stopwords("pt"),
+                       "tôda","tambêm","ha","póde","afinal")) %>%
+  mutate( stm_word2 = tm::stemDocument(word2, "portuguese")) %>% 
+  mutate( stm_word2 = tm::stemCompletion(stm_word2, word2, type = "prevalent")) %>% 
+  count(word1, stm_word2, sort=T) %>% 
+  spread( word1, n, fill=0) %>% 
+  mutate(
+    total=ele+ela,
+    ele = (ele+1)/sum(ele+1),
+    ela = (ela+1)/sum(ela+1),
+    log_ratio = log2(ela/ele),
+    abs_ratio = abs(log_ratio)
+  ) %>% 
+  arrange(desc(log_ratio)) %>% 
+  top_n(15, abs_ratio) %>% 
+  mutate(stm_word2=fct_reorder(stm_word2, log_ratio)) %>% 
+  ggplot(aes(x=stm_word2, y=log_ratio, fill=log_ratio<0)) +
+  geom_col() + 
+  coord_flip() +
+  theme_minimal()
+
+tm::stopwords("pt")
+tm::removeNumbers()
+tm::removePunctuation()
+tm::stemDocument("portuguese")

### 2021-11-16 ele_ela_gutenberg/book_text_import.R

@@ -1,0 +1,51 @@
+library(tidyverse)
+
+books <- readRDS("./ele_ela_gutenberg/books.rds")
+book_files <- readRDS("./ele_ela_gutenberg/books_file.rds")
+
+bks <- books %>% 
+  select(book_id, author, title, subject, language) %>% 
+  unnest(c(author, title, subject, language)) %>% 
+  distinct() %>% 
+  # filter( str_detect(str_to_lower(subject), "fiction"),
+  #         language == "Portuguese" ) %>% 
+  inner_join(book_files, by="book_id")
+
+book_lines <- bks %>% 
+  split(1:nrow(.)) %>% 
+  map_df(function(.book){
+
+    raw_text <- readLines(.book$filename, encoding = "UTF-8")
+    
+    text_lines <- raw_text %>% 
+      str_c(" ") %>% 
+      str_replace("^ $","<empty_line>") %>% 
+      str_c(collapse = "") %>% 
+      str_split("<empty_line>") %>% 
+      unlist() %>% 
+      str_trim()
+    
+    text_lines <- text_lines[text_lines!=""]
+    
+    tibble(line_text = text_lines) %>% 
+      mutate(
+        book_id = .book$book_id,
+        lines = length(text_lines),
+        line_number = row_number()
+      ) %>% 
+      nest(text_lines=c(line_number, line_text)) %>% 
+      return()
+    
+    # .book %>%
+    #   select(book_id) %>% 
+    #   mutate( text = list(text_lines),
+    #           lines = length(text_lines)) %>% 
+    #   unnest(text) %>% 
+    #   mutate( line_number = row_number() ) %>% 
+    #   nest(text_lines=c(line_number, text)) %>% 
+    #   return()
+    
+  })
+
+book_lines %>% 
+  saveRDS("./ele_ela_gutenberg/book_text.rds")

### 2021-11-16 ele_ela_gutenberg/gender_description.R

@@ -1,0 +1,43 @@
+library(tidyverse)
+library(tidytext)
+
+books <- readRDS("./ele_ela_gutenberg/book_text.rds")
+
+trigrams <- books %>% 
+  unnest(text_lines) %>% 
+  mutate(line_text = str_to_lower(line_text)) %>% 
+  unnest_tokens(bigram, line_text, token = "ngrams", n=3, collapse = F)
+
+trigrams_separated <- trigrams %>% 
+  separate(bigram, c("gender","verb","description"), sep=" ")
+
+saveRDS(trigrams_separated,"./ele_ela_gutenberg/book_trigrams.rds")
+
+he_she_words <- trigrams_separated %>% 
+  filter( gender %in% c("ele","ela", "eles","elas","elle","ella","elles","elllas") ) %>% 
+  mutate( gender = str_remove(gender,"s") ) %>% 
+  mutate( gender = str_replace(gender,"ll","l") ) %>% 
+  filter( verb %in% c("é","eram", "são", "serão"))
+
+he_she_words %>% 
+  filter(!description %in% c(tm::stopwords("pt"),
+                       "tôda","tambêm","ha","póde","afinal",
+                       "elle","ella")) %>%
+  mutate( stm_description = tm::stemDocument(description, "portuguese")) %>% 
+  mutate( stm_description = tm::stemCompletion(stm_description, description, type = "prevalent")) %>% 
+  count(gender, stm_description, sort=T) %>% 
+  spread( gender, n, fill=0) %>% 
+  mutate(
+    total=ele+ela,
+    ele = (ele+1)/sum(ele+1),
+    ela = (ela+1)/sum(ela+1),
+    log_ratio = log2(ela/ele),
+    abs_ratio = abs(log_ratio)
+  ) %>% 
+  arrange(desc(log_ratio)) %>% 
+  top_n(18, abs_ratio) %>% 
+  mutate(stm_description=fct_reorder(stm_description, log_ratio)) %>% 
+  ggplot(aes(x=stm_description, y=log_ratio, fill=log_ratio<0)) +
+  geom_col() + 
+  coord_flip() +
+  theme_minimal()

### 2021-11-16 fitbit_cpap/eda_sleepscore.R

@@ -1,0 +1,64 @@
+library(tidyverse)
+library(lubridate)
+library(janitor)
+
+setwd("C:/Projects/r-x-ii/fitbit_cpap")
+
+sh <- read_csv("./data/raw/sleepyhead/SleepyHead_Username_Summary_2018-01-16_2020-08-16.csv") %>% 
+  janitor::clean_names() %>% 
+  set_names(paste0("cpap_", names(.)))
+
+
+sc <- read_csv("./data/raw/sleep-score/sleep_score.csv") %>% 
+  janitor::clean_names() %>% 
+  set_names(paste0("fitbit_",names(.)))
+
+glimpse(sh)
+glimpse(sc)
+
+sh %>% 
+  count(cpap_date, sort = T)
+
+sc %>% 
+  mutate(fitbit_date = floor_date(fitbit_timestamp, unit = "days")) %>% 
+  count(fitbit_date, sort=T)
+
+sleep_df <- sc %>% 
+  mutate(fitbit_date = floor_date(fitbit_timestamp, unit = "days")) %>% 
+  inner_join(sh, by=c("fitbit_date"="cpap_date")) %>% 
+  select(-fitbit_sleep_log_entry_id,-fitbit_timestamp, -fitbit_date, -cpap_start, -cpap_end) %>%
+  mutate( cpap_total_time=as.numeric(cpap_total_time)/(60*60) ) %>% 
+  # Excellent: 90-100
+  # Good: 80-89
+  # Fair: 60-79
+  # Poor: Less than 60  
+  mutate( fitbit_sleep_class = case_when(
+    fitbit_overall_score >= 90 ~ "excellent",
+    fitbit_overall_score >= 80 ~ "good",
+    fitbit_overall_score >= 60 ~ "fair",
+    fitbit_overall_score <  60 ~ "poor",
+  )) %>% 
+  mutate( fitbit_sleep_class = factor(fitbit_sleep_class, 
+                                      levels=c("poor","fair","good","excellent"), 
+                                      ordered=T )) %>% 
+  select(-caret::nearZeroVar(.))
+
+
+install.packages("summarytools")
+install.packages("DataExplorer")
+
+library(summarytools)
+library(skimr)
+
+skimr::skim(sleep_df)
+
+freq(sleep_df, style="rmarkdown")
+ctable(sleep_df, method="render")
+dfSummary(sleep_df)
+
+
+browser(dfSummary(sleep_df))
+
+sleep_df %>% 
[…]

### 2021-11-16 gugachacra/import_twitter.R

@@ -1,0 +1,57 @@
+# install.packages("twitteR")
+library(yaml)
+library(tidyverse)
+library(rtweet)
+library(tidytext)
+
+## store api keys (these are fake example values; replace with your own keys)
+cfg <- read_yaml("./gugachacra/twitter.yml")
+
+## authenticate via web browser
+token <- create_token(
+  consumer_key = cfg$apiKey,
+  consumer_secret = cfg$apiSecret,
+  access_token = cfg$accessToken,
+  access_secret = cfg$accessTokenSecret
+)
+
+# rtm <- search_tweets(
+#   "@gugachacra", n = 5000, include_rts = FALSE
+# )
+
+rt <- get_timeline("gugachacra", n = 5000)
+
+dtm <- rt %>% 
+  select(status_id, text) %>% 
+  # mutate( text = str_remove_all(text,
+  # "https?:\\/\\/(www\\.)?[-a-zA-Z0-9@:%._\\+~#=]{1,256}\\.[a-zA-Z0-9()]{1,6}\\b([-a-zA-Z0-9()@:%_\\+.~#?&//=]*)")) %>%
+  # # mutate( text_stm = tm::stemCompletion(text_stm, text, "prevalent"))
+  mutate( text = tm::removeNumbers(text) ) %>%
+  mutate( text = tm::removePunctuation(text) ) %>% 
+  unnest_tokens(word, text) %>% 
+  anti_join(get_stopwords("pt"), by="word") %>%
+  anti_join(get_stopwords(), by="word") %>%
+  # mutate( word = tm::stemDocument(word, "portuguese")) %>% 
+  select(status_id, word) %>% 
+  filter( word != "" ) %>% 
+  add_count(status_id, word) %>%
+  cast_dtm(document = status_id, term = word, value = n)
+
+library(topicmodels)
+
+lda_model <- LDA(dtm, 5)
+
+library(broom)
+
+word_topics <- tidy(lda_model)
+
+word_topics %>% 
+  group_by(topic) %>% 
+  top_n(8, beta) %>% 
+  ungroup() %>% 
+  arrange(topic, -beta) %>% 
+  mutate( term = reorder_within(term, beta, topic)) %>% 
+  ggplot(aes(beta, term, fill=factor(topic))) +
+    geom_col(show.legend = F) +
+    facet_wrap(~topic, scales="free") +
+    scale_y_reordered()

### 2021-11-16 queimadas/queimadas_eda.R

@@ -1,0 +1,58 @@
+library(tidyverse)
+library(lubridate)
+library(gganimate)
+
+# dados DB queimadas
+dbq_raw <- read_csv("./queimadas/Focos_2020-01-01_2020-09-24.csv")
+
+dbq <- dbq_raw %>% 
+  mutate(datahora = ymd_hms(datahora))
+
+dbq %>% 
+  mutate( dia = floor_date(datahora, unit="days" ) ) %>% 
+  filter(datahora >= ymd("20200801")) %>% 
+  filter(dia == ymd(20200908)) %>% 
+  glimpse()
+
+# Além de monitorar a localização dos focos de calor, os satélites registram a intensidade da queimada.
+# Medida como FRP (sigla em inglês para potência radiativa do fogo), essa variável apresentou 
+# uma média alta em julho e, em agosto, sofreu uma escalada ainda mais acentuada.
+
+# quais satelites tem frp?
+dbq %>% 
+  filter(!is.na(frp)) %>% 
+  select(satelite) %>% 
+  distinct()
+
+# quais satelites com frp tem mais registros
+dbq %>% 
+  filter(!is.na(frp)) %>% 
+  count(satelite, sort=T)
+
+# tem os mesmos periodos e região?
+dbq %>% 
+  filter(!is.na(frp)) %>% 
+  group_by(satelite) %>% 
+  summarise(
+    min_date = min(datahora), 
+    max_date = max(datahora), 
+    min_lat  = min(latitude), 
+    max_lat  = max(latitude), 
+    min_lon  = min(longitude), 
+    max_lon  = max(longitude),
+    registros = n()
+  )
+
+g <- dbq %>% 
+  filter(satelite=="NPP-375",
+         datahora %within% interval(ymd(20200915), ymd(20200916))) %>% 
+  ggplot(aes(x=longitude, y=latitude, group=datahora)) +
+  geom_point(aes(color=frp)) +
+  scale_color_gradient(low="darkred",high="yellow") +
+  theme_minimal()
+
+g + 
+  transition_time(datahora) +
+  labs(title="Fire in Brazil",
+       subtitle = "Date:  {frame_time} - source: IBAMA") +
+  shadow_wake(wake_length = .05)

### 2021-11-16 tidymodels/apm_segmentation.R

@@ -1,0 +1,46 @@
+# install.packages("AppliedPredictiveModeling")
+library(AppliedPredictiveModeling)
+library(tidyverse)
+library(tidymodels)
+library(skimr)
+library(corrplot)
+
+## EDA
+
+# dataset
+data(segmentationOriginal)
+
+# overview
+str(segmentationOriginal)
+skim(segmentationOriginal)
+
+segData <- segmentationOriginal %>% 
+  # removing #ID, #Train/test, #other feature flags (status)
+  select(-Cell, -Case, -contains("Status") )
+
+# corr
+
+## Data Preparation
+
+# training/testing
+seg_split <- initial_split(segData, prop = .8, strata = Class)
+seg_train <- training(seg_split)
+seg_test  <- testing(seg_split)
+
+# transform
+seg_rec <- recipe(Class~., data=seg_train) %>% 
+  step_zv(all_predictors()) %>% 
+  step_nzv(all_predictors())  %>% 
+  step_BoxCox(all_predictors()) %>% 
+  step_center(all_predictors()) %>% 
+  step_scale(all_predictors()) %>% 
+  step_pca(all_predictors())
+
+seg_rec_prep <- prep(seg_rec)
+  
+juice(seg_rec_prep)
+
+step_corr(all_predictors())
+
+step_nzv()  
+
